---
title: "Example 3.4"
subtitle: "Disease mapping: from foundations to multidimensional modeling"
author: Martinez-Beneito M.A. and Botella-Rocamora P.
output: pdf_document
editor_options: 
  chunk_output_type: console
---

This document reproduces the analysis made at Example 3.4 of the book: "Disease mapping: from foundations to multidimensional modeling" by Martinez-Beneito M.A. and Botella-Rocamora P., published by CRC press in 2019. You can watch the analysis made with full detail at this pdf document, or even execute it if you want with the material available at <https://github.com/MigueBeneito/DMBook>. Anyway, this pdf file should be enough for following most of the details of the analysis made for this example.

The statistical analysis below has been run in `R`, by additionally using the library `Rmarkdown`, so be sure that you have this software installed if you want to reproduce by yourself the content of this document. In that case we advise you to download first the annex material at <https://github.com/MigueBeneito/DMBook>, open with `Rstudio` the corresponding `.Rproj` file that you will find at the folder corresponding to this example and compile the corresponding `.Rmd` document. This will allow you to reproduce the whole statistical analysis below.
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, width = 80, tidy = TRUE, tidy.opts = list(width.cutoff = 70))
set.seed(1)
```

## Libraries and data loading
```{r libraries and data loading, warning=FALSE, message=FALSE}
#Libraries loading
#-----------------
if(!require(R2WinBUGS)){install.packages("R2WinBUGS");library(R2WinBUGS)}
if(!require(INLA)){install.packages("INLA",repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE);library(INLA)}
if(!require(pbugs)){
    if(!require(devtools)){
        install.packages("devtools");devtools::install_github("fisabio/pbugs")
    }
    else{
        install_github("fisabio/pbugs")
    }
}

#Data loading
#------------
load("../Data/OralCancerTimeTrends.RData")
```

## Data preparation
```{r data preparation}
#covariate
year = 1991:2011
year.centered = year - mean(year)

#rates
rate = 1e+05 * O/Pop

```

## INLA call to splines model with default priors
```{r INLA call with default priors, warning=FALSE}
# basis of functions of 9 elements
base9 = matrix(nrow = 21, ncol = 9)
for (j in 1:9) {
    base9[, j] = dnorm(1:21, 1 + j * 2, 2)
}

# INLA call
data = data.frame(rate = rate, year = year - 2001, id.node = 1:21)
form1 = rate ~ year + f(id.node, model = "z", Z = base9)
result1 = inla(form1, data = data, control.compute = list(dic = TRUE))

# results summary
summary(result1)
```

## WinBUGS call to random effects splines model
```{r WinBUGS splines model}
#WinBUGS model
model.random = function(){
	for(i in 1:n){
		rate[i] ~ dnorm(mu[i],tau.rate)
		mu[i] <- beta[1]+beta[2]*year[i]+inprod2(base[i,],gamma[])
	}
	for(i in 1:nBase){
		gamma[i] ~ dnorm(0,tau.random)
	}
	beta[1] ~ dflat()
	beta[2] ~ dflat()

	tau.rate <- pow(sd.rate,-2)
	tau.random <- pow(sd.random,-2)
	sd.rate ~ dunif(0,10)
	sd.random ~ dunif(0,10)
}

# WinBUGS run with a basis of 9 functions
dataWB = list(rate = rate, year = year.centered, base = base9, nBase = 9, n = 21)
inits = function() {
    list(beta = rnorm(2), gamma = rnorm(9))
}
param = c("beta", "gamma", "sd.rate", "sd.random", "mu")

time.bugs = system.time(ResulWB <- bugs(data = dataWB, inits = inits, param = param, model = model.random, n.iter = 10000, n.burnin = 1000, n.thin = 9, DIC = FALSE, bugs.seed = 1))

#We call WinBUGS twice, with bugs and pbugs functions, for comparing the computational performance of inla and WinBUGS
time.pbugs = system.time(ResulWB <- pbugs(data = dataWB, inits = inits, param = param, model = model.random, n.iter = 10000, n.burnin = 1000, n.thin = 9, DIC = FALSE, bugs.seed = 1))
```

## Alternative INLA models
```{r, warning=FALSE}
# Gamma(2,0.00005) prior on the precision parameter of the splines coefficients
form1.b = rate ~ year + f(id.node, model = "z", Z = base9, hyper = list(prec = list(param = c(2, 5e-05))))
result1.b = inla(form1.b, data = data, control.compute = list(dic = TRUE))
summary(result1.b)

# Uniform prior on the standard deviation parameter of the splines coefficients
sdunif = "expression:
    logdens=-log_precision/2;
    return(logdens)"
formula2 = rate ~ year + f(id.node, model = "z", Z = base9, hyper = list(prec = list(prior = sdunif)))
result2 <- inla(formula2, data = data, control.compute = list(dic = TRUE), control.family = list(hyper = list(prec = list(prior = sdunif))))
summary(result2)

# RW1 prior (with uniform prior on the standard deviation) for the year effect
formula3 = rate ~ f(id.node, model = "rw1", hyper = list(theta = list(prior = sdunif)))
result3 <- inla(formula3, data = data, control.compute = list(dic = TRUE), control.family = list(hyper = list(theta = list(prior = sdunif))))
summary(result3)

# RW2 prior (with uniform prior on the standard deviation) for the year effect
formula4 = rate ~ f(id.node, model = "rw2", hyper = list(theta = list(prior = sdunif)))
result4 <- inla(formula4, data = data, control.compute = list(dic = TRUE), control.family = list(hyper = list(theta = list(prior = sdunif))))
summary(result4)
```

##Computational comparison between `INLA` and `WinBUGS`
```{r}
result2$cpu.used
time.bugs
time.pbugs
```

##DIC comparison between several INLA models
```{r}
#splines model
result2$dic$dic
#RW1
result3$dic$dic
#RW2
result4$dic$dic
```

# Plot illustrating the different fits made
```{r}
par(mfrow = c(1, 2))
plot(year, rate)
lines(year, ResulWB$mean$mu)
lines(year, result2$summary.fitted.values[, 1], lty = 2)
lines(year, result1$summary.fitted.values[, 1], lty = 3)
legend(1999, 9.7, lty = c(3, 1, 2), legend = c("INLA (Gamma prior)", "WinBUGS (Uniform prior)", "INLA (Uniform prior)"), cex = 0.8)
title("Gaussian basis modelling")
plot(year, rate)
lines(year, result2$summary.fitted.values[, 1], lty = 2)
lines(year, result3$summary.fitted.values[, 1], lty = 1)
lines(year, result4$summary.fitted.values[, 1], lty = 3)
legend(2003, 9.7, lty = c(2, 1, 3), legend = c("Gaussian basis", "RW1", "RW2"), cex = 0.8)
title("Alternative structures")
```
