---
title: "Example 5.2"
subtitle: "Disease mapping: from foundations to multidimensional modeling"
author: Martinez-Beneito M.A. and Botella-Rocamora P.
output: pdf_document
editor_options: 
  chunk_output_type: console
---

This document reproduces the analysis made at Example 5.2 of the book: "Disease mapping: from foundations to multidimensional modeling" by Martinez-Beneito M.A. and Botella-Rocamora P., published by CRC press in 2019. You can watch the analysis made with full detail at this pdf document, or even execute it if you want with the material available at <https://github.com/MigueBeneito/DMBook>. Anyway, this pdf file should be enough for following most of the details of the analysis made for this example.

The statistical analysis below has been run in `R`, by additionally using the library `Rmarkdown`, so be sure that you have this software installed if you want to reproduce by yourself the content of this document. In that case we advise you to download first the annex material at <https://github.com/MigueBeneito/DMBook>, open with `Rstudio` the corresponding `.Rproj` file that you will find at the folder corresponding to this example and compile the corresponding `.Rmd` document. This will allow you to reproduce the whole statistical analysis below.
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, width = 80, tidy = TRUE, tidy.opts = list(width.cutoff = 70))
set.seed(1)
```

This document has been executed with real data that are not provided in order to preserve their confidentiality. Slightly modified data are provided instead, as described in Chapter 1 of the book. Thus, when reproducing this document you will not obtain exactly the same results, although they should be very close to those shown here.

##Libraries and data loading
```{r libraries and data loading, warning=FALSE, message=FALSE}
#Libraries loading
#-----------------
if(!require(RColorBrewer)){install.packages("RColorBrewer");library(RColorBrewer)}
if(!require(rgdal)){install.packages("rgdal");library(rgdal)}
if(!require(spdep)){install.packages("spdep");library(spdep)}
if(!require(pbugs)){
    if(!require(devtools)){
        install.packages("devtools");devtools::install_github("fisabio/pbugs")
    }
    else{
        install_github("fisabio/pbugs")
    }
}

#Data loading
#------------
# For reproducing the document, the following line should be changed to load("../Data/AIDS-mod.Rdata") since that file contains the modified data making it possible to reproduce this document.
load("../Data/AIDS.Rdata")
# The data loaded contain the following objects:
# carto.val: SpatialPolygonsDataFrame with the census tracts of Valencia city
# Deprivation: Deprivation index for the census tracts of Valencia city
# Obs.AIDS: Observed cases per census tract
# Exp.AIDS: Observed cases per census tract
```

##`R` function for calculating the DIC criterion of the models fitted
The function below computes the DIC criterion for disease mapping models fitted with `WinBUGS`. It returns DIC values comparable to those reported by `INLA`, in contrast to `WinBUGS`. See annex material for Example 4.3.
```{r}
# Arguments: Simu.sSMRs: matrix of dimensions n.IterXn.Units where n.Iter are the number of MCMC iterations saved and n.Units the number of spatial units in the analysis. You will typically find this
# as a submatrix of the sims.matrix element of any bugs object. O: Vector of length n.Units with the observed deaths per spatial unit. E: Vector of length n.Units with the expected deaths per spatial
# unit.
DICPoisson = function(Simu.sSMRs, O, E) {
    mu = t(apply(Simu.sSMRs/100, 1, function(x) {
        x * E
    }))
    D = apply(mu, 1, function(x) {
        -2 * sum(O * log(x) - x - lfactorial(O))
    })
    Dmean = mean(D)
    mumean = apply(Simu.sSMRs/100, 2, mean) * E
    DinMean = -2 * sum(O * log(mumean) - mumean - lfactorial(O))
    # if(save==TRUE){return(c(Dmedia,Dmedia-DenMedia,2*Dmedia-DenMedia))}
    cat("D=", Dmean, "pD=", Dmean - DinMean, "DIC=", 2 * Dmean - DinMean, "\n")
}
```

## Data management
```{r}
# Neighbouring structure for the Valencia city census tracts
carto.val.nb = poly2nb(carto.val, snap = 1)
carto.val.wb = nb2WB(carto.val.nb)

# Categorization of the deprivation index in 5 and 25 groups
cuts.5 <- quantile(Deprivation, prob = c(0, 0.2, 0.4, 0.6, 0.8, 1))
Deprivation.5 <- as.numeric(cut(Deprivation, cuts.5, include.lowest = TRUE))
cuts.25 <- quantile(Deprivation, prob = (0:25/25))
Deprivation.25 <- as.numeric(cut(Deprivation, cuts.25, include.lowest = TRUE))
```

## Ecological regression with linear term
```{r}
RegEcoLinear = function(){
  for(i in 1:n){
    O[i] ~ dpois(lambda[i])
    log(lambda[i]) <- log(E[i])+log.theta[i]
    log.theta[i] <- mu+beta*Deprivation[i]+sd.sp*sp[i]+sd.h*het[i]
    sSMR.withoutDep[i] <- 100*exp(mu+sd.sp*sp[i]+sd.h*het[i])
    sSMR.withDep[i] <- 100*exp(log.theta[i])
    het[i] ~ dnorm(0,1)
  }
  sp[1:n] ~ car.normal(adj[],weights[],num[],1)  
  
  for (i in 1:nVec){
    weights[i] <- 1
  }
  
  #Prior distributions
  mu ~ dflat()
  sd.sp ~ dunif(0,5)
  sd.h ~ dunif(0,5)
  beta ~ dflat()
}

data = list(O = Obs.AIDS, E = Exp.AIDS, Deprivation = Deprivation, n = length(Obs.AIDS), nVec = length(carto.val.wb$adj), num = carto.val.wb$num, adj = carto.val.wb$adj)
inits = function() {
    list(mu = rnorm(1), beta = rnorm(1), sd.sp = runif(1), sd.h = runif(1), sp = rnorm(length(Obs.AIDS)), het = rnorm(length(Obs.AIDS)))
}
param = c("sSMR.withDep", "sSMR.withoutDep", "mu", "beta", "sd.sp", "sd.h")
ResulLin = pbugs(data = data, inits = inits, parameters = param, model.file = RegEcoLinear, n.iter = 10000, n.burnin = 1000, DIC = F, bugs.seed = 1)
# Computing time
ResulLin$exec_time
# Convergence checking
summary(ResulLin$summary[, "Rhat"])
summary(ResulLin$summary[, "n.eff"])
# Results summaries
round(ResulLin$summary["beta", ], 3)
exp(ResulLin$mean$beta * (max(Deprivation) - min(Deprivation)))
```


## Ecological regression with 5 independent categories
```{r}
RegEco5Cat = function(){
  for(i in 1:n){
    O[i] ~ dpois(lambda[i])
    log(lambda[i]) <- log(E[i])+log.theta[i]
    log.theta[i] <- mu+beta[Deprivation.5[i]]+sd.sp*sp[i]+sd.h*het[i]
    sSMR.withoutDep[i] <- 100*exp(mu+sd.sp*sp[i]+sd.h*het[i])
    sSMR.withDep[i] <- 100*exp(log.theta[i])
    het[i] ~ dnorm(0,1)
  }
  sp[1:n] ~ car.normal(adj[],weights[],num[],1)  
  
  for (i in 1:nVec){
    weights[i] <- 1
  }
  
  #Prior distributions
  mu ~ dflat()
  sd.sp ~ dunif(0,5)
  sd.h ~ dunif(0,5)
  beta[1] <- -sum(beta[2:5])
  for(i in 2:5){
    beta[i] ~ dflat()
  }
}

data = list(O = Obs.AIDS, E = Exp.AIDS, Deprivation.5 = Deprivation.5, n = length(Obs.AIDS), nVec = length(carto.val.wb$adj), num = carto.val.wb$num, adj = carto.val.wb$adj)
inits <- function() {
    list(mu = rnorm(1), beta = c(NA, rnorm(4)), sd.sp = runif(1), sd.h = runif(1), sp = rnorm(length(Obs.AIDS)), het = rnorm(length(Obs.AIDS)))
}
param = c("sSMR.withDep", "sSMR.withoutDep", "mu", "beta", "sd.sp", "sd.h")
Resul5Cat = pbugs(data = data, inits = inits, parameters = param, model.file = RegEco5Cat, n.iter = 10000, n.burnin = 1000, DIC = F, bugs.seed = 1)
# Computing time
Resul5Cat$exec_time
# Convergence checking
summary(Resul5Cat$summary[, "Rhat"])
summary(Resul5Cat$summary[, "n.eff"])
```

## Ecological regression with 25 independent categories
```{r}
RegEco25Cat = function(){
  for(i in 1:n){
    O[i] ~ dpois(lambda[i])
    log(lambda[i]) <- log(E[i])+log.theta[i]
    log.theta[i] <- mu+sd.beta*beta[Deprivation.25[i]]+sd.sp*sp[i]+sd.h*het[i]
    sSMR.withoutDep[i] <- 100*exp(mu+sd.sp*sp[i]+sd.h*het[i])
    sSMR.withDep[i] <- 100*exp(log.theta[i])
    het[i] ~ dnorm(0,1)
  }
  sp[1:n] ~ car.normal(adj[],weights[],num[],1)  
  
  for (i in 1:nVec){
    weights[i] <- 1
  }
  
  beta[1:25] ~ car.normal(adj.beta[],weights.beta[],num.beta[],1)
  for (i in 1:nVec.beta){
    weights.beta[i] <- 1
  }
  
  # Prior distributions
  mu ~ dflat()
  sd.sp ~ dunif(0,5)
  sd.h ~ dunif(0,5)
  sd.beta ~ dunif(0,5)
}

num.beta = c(1, rep(2, 23), 1)
adj.beta = c(2)
for (i in 2:24) {
    adj.beta = c(adj.beta, i - 1, i + 1)
}
adj.beta = c(adj.beta, 24)

data = list(O = Obs.AIDS, E = Exp.AIDS, Deprivation.25 = Deprivation.25, n = length(Obs.AIDS), nVec = length(carto.val.wb$adj), num = carto.val.wb$num, adj = carto.val.wb$adj, nVec.beta = length(adj.beta), 
    num.beta = num.beta, adj.beta = adj.beta)
inits = function() {
    list(mu = rnorm(1), beta = rnorm(25), sd.sp = runif(1), sd.beta = runif(1), sd.h = runif(1), sp = rnorm(length(Obs.AIDS)), het = rnorm(length(Obs.AIDS)))
}
param = c("sSMR.withDep", "sSMR.withoutDep", "mu", "beta", "sd.sp", "sd.h", "sd.beta")
Resul25Cat = pbugs(data = data, inits = inits, parameters = param, model.file = RegEco25Cat, n.iter = 10000, n.burnin = 1000, DIC = F, bugs.seed = 1)
# Computing time
Resul25Cat$exec_time
# Convergence checking
summary(Resul25Cat$summary[, "Rhat"])
summary(Resul25Cat$summary[, "n.eff"])
```

##DIC comparison of the previous three models
```{r}
DICPoisson(ResulLin$sims.matrix[, grep("sSMR.withDep", dimnames(ResulLin$sims.matrix)[[2]])], Obs.AIDS, Exp.AIDS)
DICPoisson(Resul5Cat$sims.matrix[, grep("sSMR.withDep", dimnames(Resul5Cat$sims.matrix)[[2]])], Obs.AIDS, Exp.AIDS)
DICPoisson(Resul25Cat$sims.matrix[, grep("sSMR.withDep", dimnames(Resul25Cat$sims.matrix)[[2]])], Obs.AIDS, Exp.AIDS)
```

##Figure in the example
```{r}
par(mfrow = c(2, 2))
# top-left plot
from = range(Deprivation)[1]
to = range(Deprivation)[2]
plot(from, cuts.25[2], ylim = c(-1.5, 1.5), xlim = range(Deprivation), type = "n", ylab = "Log-risk", xlab = "Deprivation")
abline(c(ResulLin$mean$mu, ResulLin$mean$beta))

y = mean(Resul25Cat$sims.matrix[, "sd.beta"] * Resul25Cat$sims.matrix[, "beta[1]"])
segments(x0 = from, x1 = cuts.25[2], y0 = y, y1 = y, col = "blue")
for (i in 2:24) {
    beta.1 = which(dimnames(Resul25Cat$sims.matrix)[[2]] == "beta[1]")
    y = mean(Resul25Cat$sims.matrix[, "sd.beta"] * Resul25Cat$sims.matrix[, beta.1 + i - 1])
    segments(x0 = cuts.25[i], x1 = cuts.25[i + 1], y0 = y, y1 = y, col = "blue")
}
y = mean(Resul25Cat$sims.matrix[, "sd.beta"] * Resul25Cat$sims.matrix[, "beta[25]"])
segments(x0 = cuts.25[25], x1 = to, y0 = y, y1 = y, col = "blue")

segments(x0 = from, x1 = cuts.5[2], y0 = Resul5Cat$mean$beta[1], y1 = Resul5Cat$mean$beta[1], col = 2)
for (i in 2:4) {
    segments(x0 = cuts.5[i], x1 = cuts.5[i + 1], y0 = Resul5Cat$mean$beta[i], y1 = Resul5Cat$mean$beta[i], col = 2)
}
segments(x0 = cuts.5[5], x1 = to, y0 = Resul5Cat$mean$beta[5], y1 = Resul5Cat$mean$beta[5], col = 2)
legend(x = "bottomright", lty = 1, col = c(1:2, "blue"), legend = c("Linear", "5 Depr. levels", "25 Depr. levels"), inset = 0.05, cex = 0.7)

# Top-right plot
sSMR.withDep = ResulLin$mean$sSMR.withDep
sSMR.withoutDep = ResulLin$mean$sSMR.withoutDep

par(mar = c(0, 0, 1, 0) + 1)
colors = brewer.pal(7, "BrBG")[7:1]
plot(carto.val, xlim = c(721500, 736000), ylim = c(4368000, 4378000), col = colors[as.numeric(cut(sSMR.withDep, 100 * c(-0.1, 1/1.5, 1/1.25, 1/1.1, 1.1, 1.25, 1.5, 100)))])
title("sSMR", cex = 0.75)
legend(x = "bottomright", fill = colors[7:1], legend = c(">150", "125-150", "110-125", "91-110", "80-91", "66-80", "<66"), cex = 0.75, inset = 0.03)

# Bottom-left plot
plot(carto.val, xlim = c(721500, 736000), ylim = c(4368000, 4378000), col = colors[as.numeric(cut(100 * exp(ResulLin$mean$beta * Deprivation), 100 * c(-0.1, 1/1.5, 1/1.25, 1/1.1, 1.1, 1.25, 1.5, 100)))])
title("Deprivation effect", cex = 0.75)
legend(x = "bottomright", fill = colors[7:1], legend = c(">150", "125-150", "110-125", "91-110", "80-91", "66-80", "<66"), cex = 0.75, inset = 0.03)

# Bottom-right plot
plot(carto.val, xlim = c(721500, 736000), ylim = c(4368000, 4378000), col = colors[as.numeric(cut(sSMR.withoutDep, 100 * c(-0.1, 1/1.5, 1/1.25, 1/1.1, 1.1, 1.25, 1.5, 100)))])
title("Residual variability", cex = 0.75)
legend(x = "bottomright", fill = colors[7:1], legend = c(">150", "125-150", "110-125", "91-110", "80-91", "66-80", "<66"), cex = 0.75, inset = 0.03)
```

##Variance decomposition 
```{r}
# Variance explained by the deprivation component
var.deprivation = mean(sapply(ResulLin$sims.list$beta, function(x) {
    var(x * Deprivation)
}))
# Variance explained by the random effects
var.residual = mean(apply(log(ResulLin$sims.list$sSMR.withoutDep), 1, var))
# Percentage of variance explained by deprivation
c(sqrt(var.deprivation), sqrt(var.residual), var.deprivation/(var.deprivation + var.residual))
```


