---
title: "Example 6.1"
subtitle: "Disease mapping: from foundations to multidimensional modeling"
author: Martinez-Beneito M.A. and Botella-Rocamora P.
output: 
  pdf_document:
    fig_crop: false
editor_options: 
  chunk_output_type: console
---

This document reproduces the analysis made at Example 6.1 of the book: "Disease mapping: from foundations to multidimensional modeling" by Martinez-Beneito M.A. and Botella-Rocamora P., published by CRC press in 2019. You can watch the analysis made with full detail at this pdf document, or even execute it if you want with the material available at <https://github.com/MigueBeneito/DMBook>. Anyway, this pdf file should be enough for following most of the details of the analysis made for this example.

The statistical analysis below has been run in `R`, by additionally using the library `Rmarkdown`, so be sure that you have this software installed if you want to reproduce by yourself the content of this document. In that case we advise you to download first the annex material at <https://github.com/MigueBeneito/DMBook>, open with `Rstudio` the corresponding `.Rproj` file that you will find at the folder corresponding to this example and compile the corresponding `.Rmd` document. This will allow you to reproduce the whole statistical analysis below.
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, width = 80, tidy = TRUE, tidy.opts = list(width.cutoff = 70))
set.seed(1)
```

This document has been executed with real data that are not provided in order to preserve their confidentiality. Slightly modified data are provided instead, as described in Chapter 1 of the book. Thus, when reproducing this document you will not obtain exactly the same results, although they should be very close to those shown here.

##Libraries and data loading
```{r libraries and data loading, warning=FALSE, message=FALSE}
#Libraries loading
#-----------------
if(!require(RColorBrewer)){install.packages("RColorBrewer");library(RColorBrewer)}
if(!require(rgdal)){install.packages("rgdal");library(rgdal)}
if(!require(INLA)){install.packages("INLA",repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE);library(INLA)}
if(!require(splines)){install.packages("splines");library(splines)}

#Data loading
#------------
#For reproducing the document, the following line should be changed to load("../Data/ObsOral-mod.Rdata") since that file contains the modified data making it possible to reproduce this document.
load("../Data/ObsOral.Rdata")
#load("../Data/ObsOral-mod.Rdata")
load("../Data/ExpOral.Rdata")
load("../Data/VR.Rdata")
```

##`R` function for calculating the DIC criterion of the models fitted
The function below computes the DIC criterion for disease mapping models fitted with `WinBUGS`. It returns DIC values comparable to those reported by `INLA`, in contrast to `WinBUGS`. See annex material for Example 4.3.
```{r DIC}
# Arguments: Simu.sSMRs: matrix of dimensions n.IterXn.Units where n.Iter are the number of MCMC iterations saved and n.Units the number of spatial units in the analysis. You will typically find this
# as a submatrix of the sims.matrix element of any bugs object. O: Vector of length n.Units with the observed deaths per spatial unit. E: Vector of length n.Units with the expected deaths per spatial
# unit.
DICPoisson = function(Simu.sSMRs, O, E) {
    mu = t(apply(Simu.sSMRs/100, 1, function(x) {
        x * E
    }))
    D = apply(mu, 1, function(x) {
        -2 * sum(O * log(x) - x - lfactorial(O))
    })
    Dmean = mean(D)
    mumean = apply(Simu.sSMRs/100, 2, mean) * E
    DinMean = -2 * sum(O * log(mumean) - mumean - lfactorial(O))
    # if(save==TRUE){return(c(Dmedia,Dmedia-DenMedia,2*Dmedia-DenMedia))}
    cat("D=", Dmean, "pD=", Dmean - DinMean, "DIC=", 2 * Dmean - DinMean, "\n")
}
```

##Geostatistical model
```{r, warning=FALSE}
# centroids of each municipality in the Valencian Region
centroids = t(sapply(VR.cart@polygons, function(x) {
    apply(x@Polygons[[1]]@coords, 2, mean)
}))

# data for smoothing the SMRs and their location
data = data.frame(list(x = centroids[, 1], y = centroids[, 2], O = Obs.muni, E = Exp.muni))

# mesh for defining the SPDE approach on our region of study
mesh = inla.mesh.2d(loc = centroids, max.edge = c(1, 10), cutoff = 1, max.n = c(1500, 500))
plot(mesh)
points(centroids, col = "red")

# SPDE stuff
spde = inla.spde2.pcmatern(mesh, prior.range = c(0.3, 0.5), prior.sigma = c(1, 0.01))
indexs = inla.spde.make.index("s", spde$n.spde)
A = inla.spde.make.A(mesh = mesh, loc = centroids)

# Definition of a grid of points for predicting the Gaussian process to be fitted
VR.all = aggregate(VR.cart, FUN = length)
plot(VR.all)
aux = VR.all@bbox
gridx = seq(from = aux[1, 1], to = aux[1, 2], length.out = 100)
gridy = seq(from = aux[2, 1], to = aux[2, 2], length.out = 200)
grid = matrix(nrow = 0, ncol = 2)
for (i in 1:length(gridx)) {
    aux2 = apply(cbind(gridx[i], gridy), 1, function(x) {
        over(VR.all, SpatialPoints(matrix(x, ncol = 2)))
    })
    if (length(which(!is.na(aux2))) > 0) {
        grid = rbind(grid, cbind(gridx[i], gridy[which(!is.na(aux2))]))
    }
}
points(grid, pch = ".")
# 7391 points to predict
dim(grid)[1]

Ap = inla.spde.make.A(mesh = mesh, loc = grid)

# stack for estimating the geostatistical process 
stk.e = inla.stack(tag = "dat", data = list(O = Obs.muni, E = Exp.muni), A = A, effects = list(s = indexs))
# stack for predicting the geostatistical process 
stk.p = inla.stack(tag = "pred", data = list(O = NA, E = NA), A = Ap, effects = list(s = indexs))
# stk.full has stk.e and stk.p
stk.full = inla.stack(stk.e, stk.p)

formula = O ~ 1 + f(s, model = spde)
# SPDE model
resulGeo = inla(formula, family = "poisson", E = E, control.family = list(link = "log"), data = inla.stack.data(stk.e), control.compute = list(dic = TRUE), control.predictor = list(compute = TRUE, cdf = c(log(1)), link = 1, A = inla.stack.A(stk.e)))
resulGeo$cpu.used
# SPDE model with predictions
resulGeo.pred = inla(formula, family = "poisson", E = E, control.family = list(link = "log"), data = inla.stack.data(stk.full), control.compute = list(dic = TRUE), control.predictor = list(compute = TRUE, 
    cdf = c(log(1)), link = 1, A = inla.stack.A(stk.full)))
resulGeo.pred$cpu.used
```

```{r}
# summary for the main parameters in the model
summary(resulGeo)

# Choropleth maps with the sSMRs and P(sSMRs>1) at the centroid of each municipality
colors = brewer.pal(7, "BrBG")[7:1]
colorsProb = brewer.pal(7, "RdYlGn")[7:1]
cuts = c(-0.1, 1/1.5, 1/1.25, 1/1.1, 1.1, 1.25, 1.5, 100)
cuts.Prob = c(-0.1, 0.05, 0.1, 0.25, 0.75, 0.9, 0.95, 100)

sSMRs = 100 * resulGeo.pred$summary.fitted.values[1:540, "mean"]
sSMRs.P = 1 - resulGeo.pred$summary.fitted.values[1:540, "1 cdf"]

par(mfrow = c(1, 2))
par(mar = c(1, 1, 2, 1) + 0.1)
plot(VR.cart, col = colors[as.numeric(cut(sSMRs, 100 * cuts))])
title("sSMRs", cex = 0.75)
legend(x = "bottomright", fill = colors[7:1], legend = c(">150", "125-150", "110-125", "91-110", "80-91", "66-80", "<66"), cex = 0.65, inset = 0.03, title = "sSMR")
plot(VR.cart, col = colorsProb[as.numeric(cut(sSMRs.P, cuts.Prob))])
title("P(sSMRs>1)", cex = 0.75)
legend(x = "bottomright", fill = colorsProb[7:1], legend = c("0.95-1", "0.90-0.95", "0.75-0.90", "0.25-0.75", "0.10-0.25", "0.05-0.10", "0-0.05"), cex = 0.65, inset = 0.03, title = "P(sSMR>1)")
```

##Moving average (SMARS) model
```{r, message=FALSE}
source("SMARS.r")
tpo.SMARS = system.time(res.smars <- MARS(neigh = VR.nb, O = Obs.muni, E = Exp.muni, n.burnin = 5000, n.chains = 3, n.iter = 25000, n.thin = 8, jump = 10, par.h = 0.04))
# Computing time
tpo.SMARS
# Some plots to assess convergence deviance
par(mfrow = c(2, 2))
plot(res.smars$deviance[1, ], type = "l", main = "deviance")
lines(res.smars$deviance[2, ], col = 2)
lines(res.smars$deviance[3, ], col = 3)
# order of the spatial moving average process
plot(res.smars$m[1, ], type = "l", main = "order of the process")
lines(res.smars$m[2, ], col = 2)
lines(res.smars$m[3, ], col = 3)
# mean of the underlying process
plot(res.smars$mu[1, ], type = "l", main = "mean of the process")
lines(res.smars$mu[2, ], col = 2)
lines(res.smars$mu[3, ], col = 3)
# standard deviation of the underlying process
plot(res.smars$sigma[1, ], type = "l", main = "standard deviation")
lines(res.smars$sigma[2, ], col = 2)
lines(res.smars$sigma[3, ], col = 3)

# posterior distribution for m, the order of the spatial moving average process
table(res.smars$m)/prod(dim(res.smars$m))
# DIC
DICPoisson(rbind(res.smars$RR[1, , ], res.smars$RR[2, , ], res.smars$RR[3, , ]) * 100, Obs.muni, Exp.muni)
```

##Splines model
```{r}
library(splines)
# Construction of the spatial B-spline basis Cubic B-splines
p = 3
# Number of internal intervals horizontal (x1)
qh = 7
# Number of internal intervals vertical (x2)
qv = 14

# Marginal basis for longitude
x1 = coordinates(VR.cart)[, 1]
x1 = (x1 - min(x1))/(max(x1) - min(x1))
dist1 = (max(x1) - min(x1))/qh
x1l = min(x1) - dist1 * 0.05
x1r = max(x1) + dist1 * 0.05
dx1 = (x1r - x1l)/qh
knots1 = seq(x1l - p * dx1, x1r + p * dx1, by = dx1)
# The Basis for the latitudes
B1 = splineDesign(knots1, x1, p + 1)
k1 = ncol(B1)

# Marginal basis for latitude 
x2 = coordinates(VR.cart)[,2]			
x2 = (x2-min(x2))/(max(x2)-min(x2))	
dist2 = (max(x2)-min(x2))/qv
x2l = min(x2)-dist2*0.05
x2r = max(x2)+dist2*0.05
dx2 = (x2r-x2l)/qv
knots2 = seq(x2l-p*dx2, x2r+p*dx2, by=dx2)
# The basis for the longitudes
B2 = splineDesign(knots2,x2,p+1)
k2 = ncol(B2)

# Row-wise Kronecker product of the former two bases
Rten = function(X1,X2){
  one1 = matrix(1,1,ncol(X1))
  one2 = matrix(1,1,ncol(X2))
  kronecker(X1,one2)*kronecker(one1,X2)
}
# The bidimensional basis
Bs = Rten(B2,B1)
ks = ncol(Bs)


# Spatial structure matrices 
#---------------------------
order = 1

D1 = diff(diag(k1),differences=order)
P1 = t(D1)%*%D1

D2 = diff(diag(k2),differences=order)
P2 = t(D2)%*%D2

R1 = kronecker(diag(k2),P1)
R2 = kronecker(P2,diag(k1))

Cmat.s = list(inla.as.sparse(R1),inla.as.sparse(R2))


#  Define appropriate hyperprior distributions using the "expression()" function   
#	- Unif(0,Inf) for standard deviations							                
#-------------------------------------------------------------------------------
sdunif = "expression:
    logdens = -log_precision/2;
    return(logdens)"

carto.nb = VR.nb
adj = unlist(carto.nb)
# n = length(Obs.muni)

data = list(O = Obs.muni, E = Exp.muni, intercept = c(1, rep(NA, ks)), ID.area = c(NA, 1:ks))

formula = O ~ -1 + intercept + f(ID.area, model = "generic3", Cmatrix = Cmat.s, constr = TRUE, diagonal = 1e-06, hyper = list(prec1 = list(prior = sdunif), prec2 = list(prior = sdunif)))

resulSplines = inla(formula, family = "poisson", data = data, E = Exp.muni, control.predictor = list(compute = TRUE, A = cbind(rep(1, length(Obs.muni)), Bs), link = 1), control.compute = list(dic = TRUE), control.inla = list(strategy = "laplace"))
summary(resulSplines)
```

##Figure for Example 6.1
```{r}
par(mfrow = c(1, 2))
par(mar = c(0, 0, 1, 0) + 0.1)
sSMRs.pred = 100 * resulGeo.pred$summary.fitted.values[540 + 1:7391, "mean"]
plot(VR.cart, main = "Geostatistical model")
points(grid, col = colors[as.numeric(cut(sSMRs.pred, 100 * cuts))], pch = 15)
plot(VR.cart, add = TRUE)

rmes.spl = (resulSplines$summary.fitted.values[1:length(Obs.muni), 1])
plot(VR.cart, col = colors[as.numeric(cut(rmes.spl, cuts))], main = "Splines model")
```

##BYM model
```{r}
data = data.frame(O = Obs.muni, E = Exp.muni, id.node = 1:540)
# See the annex material of Example 3.6 for details on how the VR.graph file is generated
form = O ~ f(id.node, model = "bym", hyper = list(prec.spatial = list(prior = sdunif), prec.unstruct = list(prior = sdunif)), graph = "../Data/VR.graph")
resul.BYM.inla = inla(form, family = "poisson", data = data, E = E, control.compute = list(dic = TRUE))
```

##Correlations between the sSMRs of the different models
```{r}
sSMRs.Geo = sSMRs
sSMRs.SMARS = apply(res.smars$RR, 3, mean)
sSMRs.splines = (resulSplines$summary.fitted.values[1:length(Obs.muni), 1])
sSMRs.BYM = resul.BYM.inla$summary.fitted.values[, 1]
cor(cbind(sSMRs.Geo, sSMRs.SMARS, sSMRs.splines, sSMRs.BYM))
```

