---
title: "Example 7.1"
subtitle: "Disease mapping: from foundations to multidimensional modeling"
author: Martinez-Beneito M.A. and Botella-Rocamora P.
output: pdf_document
editor_options: 
  chunk_output_type: console
---
  
This document reproduces the analysis made at Example 7.1 of the book: "Disease mapping: from foundations to multidimensional modeling" by Martinez-Beneito M.A. and Botella-Rocamora P., published by CRC press in 2019. You can watch the analysis made with full detail at this pdf document, or even execute it if you want with the material available at <https://github.com/MigueBeneito/DMBook>. Anyway, this pdf file should be enough for following most of the details of the analysis made for this example.

The statistical analysis below has been run in `R`, by additionally using the library `Rmarkdown`, so be sure that you have this software installed if you want to reproduce by yourself the content of this document. In that case we advise you to download first the annex material at <https://github.com/MigueBeneito/DMBook>, open with `Rstudio` the corresponding `.Rproj` file that you will find at the folder corresponding to this example and compile the corresponding `.Rmd` document. This will allow you to reproduce the whole statistical analysis below.
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, width = 80, tidy = TRUE, tidy.opts = list(width.cutoff = 70))
set.seed(1)
```

This document has been executed with real data that are not provided in order to preserve their confidentiality. Slightly modified data are provided instead, as described in Chapter 1 of the book. Thus, when reproducing this document you will not obtain exactly the same results, although they should be very close to those shown here.

##Libraries and data loading
```{r libraries and data loading, warning=FALSE, message=FALSE}
#Libraries loading
#-----------------
if(!require(RColorBrewer)){install.packages("RColorBrewer");library(RColorBrewer)}
if(!require(rgdal)){install.packages("rgdal");library(rgdal)}
if(!require(R2WinBUGS)){install.packages("R2WinBUGS");library(R2WinBUGS)}
if(!require(INLA)){install.packages("INLA",repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE);library(INLA)}
if(!require(pbugs)){
    if(!require(devtools)){
        install.packages("devtools");devtools::install_github("fisabio/pbugs")
    }
    else{
        install_github("fisabio/pbugs")
    }
}

#Data loading
#------------
load("../Data/VR.Rdata")
load("../Data/Exp.2011.Rdata")
```

# Analysis of simulated data sets

##Generation of the simulated data sets
```{r}
nmuni = length(VR.cart)
centroids = matrix(nrow = nmuni, ncol = 2)
for (i in 1:nmuni) {
    centroids[i, ] = apply(VR.cart@polygons[[i]]@Polygons[[1]]@coords, 2, mean)
}

miny = min(centroids[, 2])
maxy = max(centroids[, 2])

colors = brewer.pal(7, "BrBG")[7:1]

# Generation of the relative risks for each setting
#-------------------------------------------------
RRisks = list()
# Setting 1
RRisks[[1]] = matrix(nrow = nmuni, ncol = 10)
for (i in 1:10) {
    RRisks[[1]][, i] = 0.5 + ((centroids[, 2] - miny)/(maxy - miny))
}

par(mfrow = c(1, 3))
par(mar = c(1, 1, 2, 1) + 0.1)
per = 1
RRisks.cut = as.numeric(cut(RRisks[[1]][, per], c(-0.1, 1/1.5, 1/1.25, 1/1.1, 1.1, 1.25, 1.5, 100)))
plot(VR.cart, col = colors[RRisks.cut], main = "Period 1")
per = 5
RRisks.cut = as.numeric(cut(RRisks[[1]][, per], c(-0.1, 1/1.5, 1/1.25, 1/1.1, 1.1, 1.25, 1.5, 100)))
plot(VR.cart, col = colors[RRisks.cut], main = "Period 5")
per = 10
RRisks.cut = as.numeric(cut(RRisks[[1]][, per], c(-0.1, 1/1.5, 1/1.25, 1/1.1, 1.1, 1.25, 1.5, 100)))
plot(VR.cart, col = colors[RRisks.cut], main = "Period 10")

# Setting 2
RRisks[[2]] = matrix(nrow = nmuni, ncol = 10)
r.per = seq(1.5, 0.5, length = 10)
for (i in 1:10) {
    RRisks[[2]][, i] = r.per[i]
}

par(mfrow = c(1, 3))
par(mar = c(1, 1, 2, 1) + 0.1)
per = 1
RRisks.cut = as.numeric(cut(RRisks[[2]][, per], c(-0.1, 1/1.5, 1/1.25, 1/1.1, 1.1, 1.25, 1.5, 100)))
plot(VR.cart, col = colors[RRisks.cut], main = "Period 1")
per = 5
RRisks.cut = as.numeric(cut(RRisks[[2]][, per], c(-0.1, 1/1.5, 1/1.25, 1/1.1, 1.1, 1.25, 1.5, 100)))
plot(VR.cart, col = colors[RRisks.cut], main = "Period 5")
per = 10
RRisks.cut = as.numeric(cut(RRisks[[2]][, per], c(-0.1, 1/1.5, 1/1.25, 1/1.1, 1.1, 1.25, 1.5, 100)))
plot(VR.cart, col = colors[RRisks.cut], main = "Period 10")

# Setting 3
RRisks[[3]] = matrix(nrow = nmuni, ncol = 10)
r.perN = seq(1.5, 0.5, length = 10)
r.perS = seq(0.5, 1.5, length = 10)

for (i in 1:10) {
    RRisks[[3]][, i] = r.perS[i] + ((centroids[, 2] - miny)/(maxy - miny)) * (r.perN[i] - r.perS[i])
}

par(mfrow = c(1, 3))
par(mar = c(1, 1, 2, 1) + 0.1)
per = 1
RRisks.cut = as.numeric(cut(RRisks[[3]][, per], c(-0.1, 1/1.5, 1/1.25, 1/1.1, 1.1, 1.25, 1.5, 100)))
plot(VR.cart, col = colors[RRisks.cut], main = "Period 1")
per = 5
RRisks.cut = as.numeric(cut(RRisks[[3]][, per], c(-0.1, 1/1.5, 1/1.25, 1/1.1, 1.1, 1.25, 1.5, 100)))
plot(VR.cart, col = colors[RRisks.cut], main = "Period 5")
per = 10
RRisks.cut = as.numeric(cut(RRisks[[3]][, per], c(-0.1, 1/1.5, 1/1.25, 1/1.1, 1.1, 1.25, 1.5, 100)))
plot(VR.cart, col = colors[RRisks.cut], main = "Period 10")

# Generation of the observed cases for each setting
Obs = list()
Obs10 = list()
for (setting in 1:3) {
    Obs[[setting]] = matrix(nrow = nmuni, ncol = 10)
    Obs10[[setting]] = matrix(nrow = nmuni, ncol = 10)
    for (i in 1:10) {
        Obs[[setting]][, i] = rpois(nmuni, Exp.2011 * RRisks[[setting]][, i])
        Obs10[[setting]][, i] = rpois(nmuni, 10 * Exp.2011 * RRisks[[setting]][, i])
    }
}
```

# Execution of the autoregressive model in `WinBUGS` for the generated data sets 
```{r}
#WinBUGS code for the autoregressive Spatio-temporal model 
Autoregressive = function(){
  for(i in 1:nmuni){
    for(j in 1:nperiods){
      Obs[i,j] ~ dpois(lambda[i,j])
      #Modelling of the mean for every municipality and period
      log(lambda[i,j]) <- log(Exp[i,j])+log.theta[i,j]
      #SMR for every municipality and period
      log.theta[i,j] <- mediainter+sd.inter*inter[j]+ST[i,j]
    }
  }
  
  #Spatio-temporal effect for the first period
  for(i in 1:nmuni){
    ST[i,1] <- pow(1-ro*ro,-0.5)*BYM[i,1]
    BYM[i,1] <- sd.het*psi[i,1]+sd.spat*phi[1,i]
    psi[i,1] ~ dnorm(0,1)
  }
  phi[1,1:nmuni] ~ car.normal(map[],w[],nvec[],1)
  #Spatio-temporal effect for the subsequent periods
  for(j in 2:nperiods){
    for(i in 1:nmuni){
      ST[i,j] <- ro*ST[i,j-1]+BYM[i,j]
      BYM[i,j] <- sd.het*psi[i,j]+sd.spat*phi[j,i]
      psi[i,j] ~ dnorm(0,1)
    }
    phi[j,1:nmuni] ~ car.normal(map[],w[],nvec[],1)
  }
  
  #Prior distribution for the mean risk for every municipality and period
  mediainter ~ dflat()
  #Prior distribution for the global time trend
  inter[1:nperiods] ~ car.normal(mapT[],wT[],nvecT[],1)
  #Prior distribution for the precision parameters in the model
  sd.inter ~ dunif(0,5)
  sd.het ~ dunif(0,5)
  sd.spat ~ dunif(0,5)
  #Prior distribution for the temporal dependence parameter
  ro ~ dunif(-1,1)
}

nperiods = 10

adjT = c(2, c(1, 3), c(2, 4), c(3, 5), c(4, 6), c(5, 7), c(6, 8), c(7, 9), c(8, 10), 9)
numT = c(1, rep(2, 8), 1)
indexT = c(1, cumsum(numT))

Result = list()
Result10 = list()

# WinBUGS calls for executing the model above on the simulated data sets
for (setting in 1:3) {
    
    Exp = matrix(nrow = nmuni, ncol = 10)
    for (j in 1:10) {
        Exp[, j] = Exp.2011
    }
    
    data = list(Obs = Obs[[setting]], Exp = Exp, nmuni = nmuni, nperiods = nperiods, w = rep(1, length(VR.wb$adj)), nvec = VR.wb$num, map = VR.wb$adj, wT = rep(1, length(adjT)), nvecT = numT, mapT = adjT)
    inits = function() {
        list(ro = runif(1, -1, 1), mediainter = rnorm(1, 0, 1), sd.inter = runif(1, 0, 0.5), sd.het = runif(1, 0, 0.5), sd.spat = runif(1, 0, 0.5), psi = matrix(rnorm(nperiods * nmuni, 0, 1), ncol = nperiods, 
            nrow = nmuni), phi = matrix(rnorm(nperiods * nmuni, 0, 1), nrow = nperiods, ncol = nmuni))
    }
    param = c("log.theta", "mediainter", "sd.inter", "sd.het", "sd.spat", "ro")
    
    Result[[setting]] = pbugs(data = data, inits = inits, parameters = param, model.file = Autoregressive, n.iter = 5000, n.burnin = 1000, DIC = F, n.chains = 3, bugs.seed = 1)
    
    data = list(Obs = Obs10[[setting]], Exp = 10 * Exp, nmuni = nmuni, nperiods = nperiods, w = rep(1, length(VR.wb$adj)), nvec = VR.wb$num, map = VR.wb$adj, wT = rep(1, length(adjT)), nvecT = numT, mapT = adjT)
    Result10[[setting]] = pbugs(data = data, inits = inits, parameters = param, model.file = Autoregressive, n.iter = 5000, n.burnin = 1000, DIC = F, n.chains = 3, bugs.seed = 1)
}
```

##Convergence checking for the models executed
```{r}
#Computing time
Result[[1]]$exec_time
#Convergence checking
summary(Result[[1]]$summary[,"Rhat"])
summary(Result[[1]]$summary[,"n.eff"])

#Computing time
Result[[2]]$exec_time
#Convergence checking
summary(Result[[2]]$summary[,"Rhat"])
summary(Result[[2]]$summary[,"n.eff"])

#Computing time
Result[[3]]$exec_time
#Convergence checking
summary(Result[[3]]$summary[,"Rhat"])
summary(Result[[3]]$summary[,"n.eff"])

#Computing time
Result10[[1]]$exec_time
#Convergence checking
summary(Result10[[1]]$summary[,"Rhat"])
summary(Result10[[1]]$summary[,"n.eff"])

#Computing time
Result10[[2]]$exec_time
#Convergence checking
summary(Result10[[2]]$summary[,"Rhat"])
summary(Result10[[2]]$summary[,"n.eff"])

#Computing time
Result10[[3]]$exec_time
#Convergence checking
summary(Result10[[3]]$summary[,"Rhat"])
summary(Result10[[3]]$summary[,"n.eff"])
```

##Variance decompositions 
```{r}
decomp = array(dim = c(3, 3, 3), dimnames = list(c("Setting 1", "Setting2", "Setting3"), c("2.5%", "50%", "97.5%"), c("S", "T", "ST")))
decomp.mean = matrix(nrow = 3, ncol = 3, dimnames = list(c("Setting1", "Setting2", "Setting3"), c("S", "T", "ST")))
for (setting in 1:3) {
    nIter = dim(Result[[setting]]$sims.list$log.theta)[1]
    components = matrix(nrow = nIter, ncol = 4)
    for (i in 1:nIter) {
        m = mean(Result[[setting]]$sims.list$log.theta[i, , ])
        S = apply(Result[[setting]]$sims.list$log.theta[i, , ], 1, mean) - m
        T = apply(Result[[setting]]$sims.list$log.theta[i, , ], 2, mean) - m
        ST = Result[[setting]]$sims.list$log.theta[i, , ] - (m + matrix(rep(S, length(T)), ncol = length(T)) + matrix(rep(T, length(S)), ncol = length(T), byrow = T))
        components[i, ] = c(m, var(S), var(T), var(as.vector(ST)))
    }
    aux = components[, c(2:4)]/apply(components[, c(2:4)], 1, sum)
    decomp[setting, , ] = apply(aux, 2, quantile, c(0.025, 0.5, 0.975))
    decomp.mean[setting, ] = apply(aux, 2, mean)
}
# quantiles of spatial, temporal and spatio-temporal components for each setting
decomp[1, , ] * 100
decomp[2, , ] * 100
decomp[3, , ] * 100
# posterior mean of the percentage of variance per component for each setting
decomp.mean * 100
```

##Variance decompositions for the data sets with expected cases 10 times larger 
```{r}
decomp10 = array(dim = c(3, 3, 3), dimnames = list(c("Setting1", "Setting2", "Setting3"), c("2.5%", "50%", "97.5%"), c("S", "T", "ST")))
decomp10.mean = matrix(nrow = 3, ncol = 3, dimnames = list(c("Setting1", "Setting2", "Setting3"), c("S", "T", "ST")))
for (setting in 1:3) {
    nIter = dim(Result10[[setting]]$sims.list$log.theta)[1]
    components = matrix(nrow = nIter, ncol = 4)
    for (i in 1:nIter) {
        m = mean(Result10[[setting]]$sims.list$log.theta[i, , ])
        S = apply(Result10[[setting]]$sims.list$log.theta[i, , ], 1, mean) - m
        T = apply(Result10[[setting]]$sims.list$log.theta[i, , ], 2, mean) - m
        ST = Result10[[setting]]$sims.list$log.theta[i, , ] - (m + matrix(rep(S, length(T)), ncol = length(T)) + matrix(rep(T, length(S)), ncol = length(T), byrow = T))
        components[i, ] = c(m, var(S), var(T), var(as.vector(ST)))
    }
    aux = components[, c(2:4)]/apply(components[, c(2:4)], 1, sum)
    decomp10[setting, , ] = apply(aux, 2, quantile, c(0.025, 0.5, 0.975))
    decomp10.mean[setting, ] = apply(aux, 2, mean)
}
# quantiles of spatial, temporal and spatio-temporal components for each setting
decomp10[1, , ] * 100
decomp10[2, , ] * 100
decomp10[3, , ] * 100
# posterior mean of the percentage of variance per component for each setting
decomp10.mean * 100
```

##Choropleth maps for period 1 and 10 for each setting
```{r}
# choropleth maps of some patterns fitted
for (setting in 1:3) {
    par(mfrow = c(1, 2))
    per = 1
    aux = cut(apply(exp(Result[[setting]]$sims.list$log.theta[, , per]), 2, mean), c(0, 0.66, 0.8, 0.9, 1.1, 1.25, 1.5, 10))
    plot(VR.cart, col = brewer.pal(7, "BrBG")[7:1][aux], main = paste0("Setting ", setting, ". Period 1."))
    per = 10
    aux = cut(apply(exp(Result[[setting]]$sims.list$log.theta[, , per]), 2, mean), c(0, 0.66, 0.8, 0.9, 1.1, 1.25, 1.5, 10))
    plot(VR.cart, col = brewer.pal(7, "BrBG")[7:1][aux], main = paste0("Setting ", setting, ". Period 10."))
}

# choropleth maps of some patterns fitted (expected and observed cases 10 times larger)
for (setting in 1:3) {
    par(mfrow = c(1, 2))
    per = 1
    aux = cut(apply(exp(Result10[[setting]]$sims.list$log.theta[, , per]), 2, mean), c(0, 0.66, 0.8, 0.9, 1.1, 1.25, 1.5, 10))
    plot(VR.cart, col = brewer.pal(7, "BrBG")[7:1][aux], main = paste0("Setting ", setting, ". Period 1."))
    per = 10
    aux = cut(apply(exp(Result10[[setting]]$sims.list$log.theta[, , per]), 2, mean), c(0, 0.66, 0.8, 0.9, 1.1, 1.25, 1.5, 10))
    plot(VR.cart, col = brewer.pal(7, "BrBG")[7:1][aux], main = paste0("Setting ", setting, ". Period 10."))
}
```

# Analysis of real data from the Valencian Region

Data not provided in order to preserve confidentiality. The model and variance decomposition carried out for the real data set is exactly the same than that used for the simulated data sets.

We reproduce here only those combination of genders and causes of death that are shown in Table 7.2.

##Execution of `WinBUGS` models for the real data sets
```{r}
load("../Data/ET-causes.Rdata")

causes = c("Men. Oral", "Men. Stomach", "Women. Stomach", "Men. Lung", "Women. Lung", "Men. Bladder", "Men. Cereb.", "Women. Cereb.", "Men. Pneumo.", "Women. Pneumo.")

nperiods.12 <- 12
adjT.12 = c(rbind(2:nperiods.12, 1:(nperiods.12 - 1)))
numT.12 = c(1, rep(2, nperiods.12 - 2), 1)
indexT.12 = c(1, cumsum(numT.12))

Result.causes = list()
for (i in 1:10) {
    data = list(Obs = Obs.Causes[[i]], Exp = Exp.Causes[[i]], nmuni = nmuni, nperiods = nperiods.12, w = rep(1, length(VR.wb$adj)), nvec = VR.wb$num, map = VR.wb$adj, wT = rep(1, length(adjT.12)), nvecT = numT.12, 
        mapT = adjT.12)
    inits = function() {
        list(ro = runif(1, -1, 1), mediainter = rnorm(1, 0, 1), sd.inter = runif(1, 0, 0.5), sd.het = runif(1, 0, 0.5), sd.spat = runif(1, 0, 0.5), psi = matrix(rnorm(nperiods.12 * nmuni, 0, 1), ncol = nperiods.12, 
            nrow = nmuni), phi = matrix(rnorm(nperiods.12 * nmuni, 0, 1), nrow = nperiods.12, ncol = nmuni))
    }
    param = c("log.theta", "mediainter", "sd.inter", "sd.het", "sd.spat", "ro")
    
    Result.causes[[i]] = pbugs(data = data, inits = inits, parameters = param, model.file = Autoregressive, n.iter = 5000, n.burnin = 1000, DIC = F, n.chains = 3, bugs.seed = 1)
}
```

##Variance decompositions
```{r}
decomp.causes = array(dim = c(10, 3, 3), dimnames = list(causes, c("2.5%", "50%", "97.5%"), c("S", "T", "ST")))
decomp.causes.mean = matrix(nrow = 10, ncol = 3, dimnames = list(causes, c("S", "T", "ST")))
for (setting in 1:10) {
    nIter = dim(Result.causes[[setting]]$sims.list$log.theta)[1]
    components = matrix(nrow = nIter, ncol = 4)
    for (i in 1:nIter) {
        m = mean(Result.causes[[setting]]$sims.list$log.theta[i, , ])
        S = apply(Result.causes[[setting]]$sims.list$log.theta[i, , ], 1, mean) - m
        T = apply(Result.causes[[setting]]$sims.list$log.theta[i, , ], 2, mean) - m
        ST = Result.causes[[setting]]$sims.list$log.theta[i, , ] - (m + matrix(rep(S, length(T)), ncol = length(T)) + matrix(rep(T, length(S)), ncol = length(T), byrow = T))
        components[i, ] <- c(m, var(S), var(T), var(as.vector(ST)))
    }
    aux = components[, c(2:4)]/apply(components[, c(2:4)], 1, sum)
    decomp.causes[setting, , ] = apply(aux, 2, quantile, c(0.025, 0.5, 0.975))
    decomp.causes.mean[setting, ] = apply(aux, 2, mean)
}
# posterior mean of the percentage of variance per component for each setting
decomp.causes.mean * 100
# posterior 2.5%, 50% and 97.5% quantiles of spatial, temporal and spatio-temporal components for each setting
decomp.causes[1, , ]
decomp.causes[2, , ]
decomp.causes[3, , ]
decomp.causes[4, , ]
decomp.causes[5, , ]
decomp.causes[6, , ]
decomp.causes[7, , ]
decomp.causes[8, , ]
decomp.causes[9, , ]
decomp.causes[10, , ]
```
