---
title: "Example 2.5"
subtitle: "Disease mapping: from foundations to multidimensional modeling"
author: Martinez-Beneito M.A. and Botella-Rocamora P.
output: pdf_document
editor_options: 
  chunk_output_type: console
---

This document reproduces the analysis made at Example 2.5 of the book: "Disease mapping: from foundations to multidimensional modeling" by Martinez-Beneito M.A. and Botella-Rocamora P., published by CRC press in 2019. You can watch the analysis made with full detail at this pdf document, or even execute it if you want with the material available at <https://github.com/MigueBeneito/DMBook>. Anyway, this pdf file should be enough for following most of the details of the analysis made for this example.

The statistical analysis below has been run in `R`, by additionally using the library `Rmarkdown`, so be sure that you have this software installed if you want to reproduce by yourself the content of this document. In that case we advise you to download first the annex material at <https://github.com/MigueBeneito/DMBook>, open with `Rstudio` the corresponding `.Rproj` file that you will find at the folder corresponding to this example and compile the corresponding `.Rmd` document. This will allow you to reproduce the whole statistical analysis below.
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, width = 80, tidy = TRUE, tidy.opts = list(width.cutoff = 70))
set.seed(1)
```

The inference carried out in this example has been undertaken with `WinBUGS`, that will be introduced in the next chapter. The details for fully understanding this code are given at Chapter 3, nevertheless you may find interesting to take it a look at this code now, or later when you read Chapter 3. 

## Libraries and data loading
```{r libraries and data loading, warning=FALSE}
#Libraries loading
#-----------------
if(!require(pbugs)){
    if(!require(devtools)){
        install.packages("devtools");devtools::install_github("fisabio/pbugs")
    }
    else{
        install_github("fisabio/pbugs")
    }
}

#Data loading
#------------
load("../Data/OralCancerTimeTrends.RData")
```

## Data preparation
```{r data preparation}
#data preparation
year = 1991:2011
year.centered = year - mean(year)

rates = 100000 * O / Pop
```

## MCMC sampling with `WinBUGS` (centered covariate)
```{r WinBUGS call (centered covariate)}
RegLin2 = function(){
	for(i in 1:n){
		rate[i] ~ dnorm(media[i],prec)
		media[i] <- beta1+beta2*year[i]
	}
	prec <- pow(sigma,-2)
	sigma ~ dunif(0,1000)
	beta1 ~ dt(0,0.000001,2)
	beta2 ~ dt(0,0.000001,2)
}

#WinBUGS call for making inference on the model above
#####################################################

# Data
data = list(year = year.centered, rate = rates, n = 21)

#Initial values    
inits = function() {
  list(beta1 = rnorm(1, 0, 10), beta2 = rnorm(1, 0, 10), sigma = runif(1, 0, 10))
}

#Parameters to save
parameters = c("beta1", "beta2", "sigma")

#WinBUGS call
RegLin2WB = pbugs(data = data, inits = inits, param = parameters, model = RegLin2, bugs.seed = 1, DIC = F,n.iter = 5500, n.burnin = 500, n.chains = 3, n.thin = 1)

#Posterior summaries
round(RegLin2WB$summary,3)
```

## MCMC sampling with `WinBUGS` (uncentered covariate)
```{r WinBUGS call (uncentered covariate)}
# Data
data = list(year = year, rate = rates, n = 21)

#Initial values    
inits = function() {
  list(beta1 = rnorm(1, 500, 20), beta2 = rnorm(1, -0.1, 0.1), sigma = runif(1, 0, 4))
}

#WinBUGS call
RegLin2WB.uncentered = pbugs(data = data, inits = inits, param = parameters, model = RegLin2, bugs.seed = 1, DIC = F,n.iter = 5500, n.burnin = 500, n.chains = 3, n.thin = 1)

#Posterior summaries
round(RegLin2WB.uncentered$summary,3)
```

## Long MCMC sampling with `WinBUGS` (uncentered covariate)
```{r Long WinBUGS run (uncentered covariate)}
#WinBUGS call
RegLin2WB.uncentered2 = pbugs(data = data, inits = inits, param = parameters, model = RegLin2, bugs.seed = 1, DIC = F,n.iter = 2000000, n.burnin = 200000, n.chains = 3, n.thin = 360)

#Posterior summaries
round(RegLin2WB.uncentered2$summary,3)
cor(RegLin2WB.uncentered2$sims.list$beta1,RegLin2WB.uncentered2$sims.list$beta2)
```

#Convergence plots
```{r Figure}
par(mfrow = c(2, 2))
plot(RegLin2WB$sims.array[, 1, 1], type = "l", main = expression(beta[1]), xlab = "Iteration", ylab = "Value")
lines(RegLin2WB$sims.array[, 2, 1], type = "l", col = 2)
lines(RegLin2WB$sims.array[, 3, 1], type = "l", col = 3)

plot(RegLin2WB.uncentered$sims.array[, 1, 1], type = "l", main = expression(beta[1]), xlab = "Iteration", ylab = "Value", ylim = c(min(RegLin2WB.uncentered$sims.array[, , 1]), max(RegLin2WB.uncentered$sims.array[, , 1])))
lines(RegLin2WB.uncentered$sims.array[, 2, 1], type = "l", col = 2)
lines(RegLin2WB.uncentered$sims.array[, 3, 1], type = "l", col = 3)

plot(RegLin2WB.uncentered2$sims.array[, 1, 1], type = "l", main = expression(beta[1]), xlab = "Iteration", ylab = "Value", ylim = c(min(RegLin2WB.uncentered2$sims.array[, , 1]), max(RegLin2WB.uncentered2$sims.array[, , 1])))
lines(RegLin2WB.uncentered2$sims.array[, 2, 1], type = "l", col = 2)
lines(RegLin2WB.uncentered2$sims.array[, 3, 1], type = "l", col = 3)

plot(RegLin2WB.uncentered2$sims.array[, 1, 2], type = "l", main = expression(beta[2]), xlab = "Iteration", ylab = "Value", ylim = c(min(RegLin2WB.uncentered2$sims.array[, , 2]), max(RegLin2WB.uncentered2$sims.array[, , 2])))
lines(RegLin2WB.uncentered2$sims.array[, 2, 2], type = "l", col = 2)
lines(RegLin2WB.uncentered2$sims.array[, 3, 2], type = "l", col = 3)
```

