---
title: "Example 9.1"
subtitle: "Disease mapping: from foundations to multidimensional modeling"
author: Martinez-Beneito M.A. and Botella-Rocamora P.
output: pdf_document
editor_options: 
  chunk_output_type: console
---

This document reproduces the analysis made at Example 9.1 of the book: "Disease mapping: from foundations to multidimensional modeling" by Martinez-Beneito M.A. and Botella-Rocamora P., published by CRC press in 2019. You can watch the analysis made with full detail at this pdf document, or even execute it if you want with the material available at <https://github.com/MigueBeneito/DMBook>. Anyway, this pdf file should be enough for following most of the details of the analysis made for this example.

The statistical analysis below has been run in `R`, by additionally using the library `Rmarkdown`, so be sure that you have this software installed if you want to reproduce by yourself the content of this document. In that case we advise you to download first the annex material at <https://github.com/MigueBeneito/DMBook>, open with `Rstudio` the corresponding `.Rproj` file that you will find at the folder corresponding to this example and compile the corresponding `.Rmd` document. This will allow you to reproduce the whole statistical analysis below.

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, width = 80, tidy = TRUE, tidy.opts = list(width.cutoff = 75))
set.seed(1)
```

This document has been executed with real data that are not provided in order to preserve their confidentiality. Slightly modified data are provided instead, as described in Chapter 1 of the book. Thus, when reproducing this document you will not obtain exactly the same results, although they should be very close to those shown here.

##Libraries and data loading
```{r libraries and data loading, warning=FALSE, message=FALSE}
#Libraries loading
#-----------------
if(!require(RColorBrewer)){install.packages("RColorBrewer");library(RColorBrewer)}
if(!require(rgdal)){install.packages("rgdal");library(rgdal)}
if(!require(pbugs)){
    if(!require(devtools)){
        install.packages("devtools");devtools::install_github("fisabio/pbugs")
    }
    else{
        install_github("fisabio/pbugs")
    }
}

#Data loading
#------------
#For reproducing the document, the following line should be changed to load("../Data/ObsMDM-mod.Rdata") since that file contains the modified data making it possible to reproduce this document.
load("../Data/ObsMDM.Rdata")
#load("../Data/ObsOral-mod.Rdata")
load("../Data/ExpMDM.Rdata")
load("../Data/VR.Rdata")
```

##`R` function for calculating the DIC criterion of the models fitted
The function below computes the DIC criterion for disease mapping models fitted with `WinBUGS`. It returns DIC values comparable to those reported by `INLA`, in contrast to `WinBUGS`. See annex material for Example 4.3.
```{r}
#Arguments: 
#Simu.sSMRs: matrix of dimensions n.IterXn.Units where n.Iter are the number of MCMC iterations saved and n.Units the number of spatial units in the analysis. You will typically find this as a submatrix of the sims.matrix element of any bugs object.
#O: Vector of length n.Units with the observed deaths per spatial unit.
#E: Vector of length n.Units with the expected deaths per spatial unit.
DICPoisson = function(Simu.sSMRs,O,E){
  mu = t(apply(Simu.sSMRs/100,1,function(x){x*E}))
  D = apply(mu,1,function(x){-2*sum(O*log(x)-x-lfactorial(O))})
  Dmean= mean(D)
  mumean = apply(Simu.sSMRs/100,2,mean)*E
  DinMean = -2*sum(O*log(mumean)-mumean-lfactorial(O))
  #if(save==TRUE){return(c(Dmedia,Dmedia-DenMedia,2*Dmedia-DenMedia))}
  cat("D=",Dmean,"pD=",Dmean-DinMean,"DIC=",2*Dmean-DinMean,"\n")
}
```

#Function monitoring the convergence of the deviance of multidimensional models
```{r}
plot.deviance <- function(resul) {
    deviances <- matrix(nrow = dim(resul$sims.array)[1], ncol = dim(resul$sims.array)[2])
    for (j in 1:(dim(resul$sims.array)[2])) {
        for (i in 1:(dim(resul$sims.array)[1])) {
            sSMR <- aperm(array(resul$sims.array[i, j, grep("sSMR", dimnames(resul$summary)[[1]])], dim = dim(Obs.md)[4:1]), 4:1)
            deviances[i, j] <- -2 * sum(dpois(as.vector(Obs.md), as.vector((sSMR/100) * Exp.md), log = T))
        }
    }
    plot(deviances[, 1], type = "l", ylim = c(min(deviances), max(deviances)), ylab = "Deviance", xlab = "Iteration saved")
    for (j in 2:(dim(resul$sims.array)[2])) {
        lines(deviances[, j], col = j)
    }
}
```

##Definition of some constants that will be extensively used for the rest of the analysis
```{r}
num.simu <- 15000
num.burnin <- 5000

Nmuni <- 540
Nper <- 5
Ndiseases <- 2
Nsex <- 2
nGroups <- c(Nmuni, Nper, Ndiseases, Nsex)  #Municipalities, Periods, Diseases, Sexes

num <- VR.wb$num
adj <- VR.wb$adj

# We define the data for the multidimensional models here since they will be common to all the models run
data <- list(Obs = Obs.md, Exp = Exp.md, nGroups = nGroups, adj = adj, num = num, C = rep(1/num, num), M = 1/num)
```

##Fully separable model
```{r}
model1.2.3.4 = function(){
  for(k in 1:(nGroups[4])){#Sex
    for(j in 1:(nGroups[3])){#Disease
      for (p in 1:(nGroups[2])){#Period
        for(i in 1:(nGroups[1])){#Municipality
          Obs[i,p,j,k] ~ dpois(lambda[i,p,j,k])
          log(lambda[i,p,j,k]) <- log(Exp[i,p,j,k])+log.theta[i,p,j,k]
          log.theta[i,p,j,k] <- mu[p,j,k]+S1234[i,p,j,k]  
          #Adds dependence between sexes
          S1234[i,p,j,k] = inprod2(S123[i,p,j,],structure4[,k])
          #Adds dependence between diseases
          S123[i,p,j,k] = inprod2(S12[i,p,,k],structure3[,j])
          #Adds dependence between periods
          S12[i,p,j,k] = inprod2(tS1[k,j,,i],structure2[,p])

          sSMR[i,p,j,k] = 100*exp(log.theta[i,p,j,k])
        }
        mu[p,j,k] = dflat()
        #Spatially dependent underlying random effects
        tS1[k,j,p,1:(nGroups[1])] = car.proper(ceros[],C[],adj[],num[],M[],1,gamma)
      }
    }
  }
  
  #Parameters for the PCAR distributions
  for(l in 1:(nGroups[1])){ceros[l] = 0}
  gamma.inf = min.bound(C[],adj[],num[],M[])
  gamma.sup = max.bound(C[],adj[],num[],M[])
  gamma = dunif(gamma.inf,gamma.sup)
  
  #Structure matrix for the second dimension (period). Structure 2 is defined as the upper Cholesky matrix of the covariance matrix of an AR1 process.
  for (pc in 1:(nGroups[2])){
    #First Row
    structure2[1,pc] = pow(ro,pc-1)*pow((1-ro*ro),-0.5)
    #Rest of columns
    for (pr in 2:(nGroups[2])){
      structure2[pr,pc] = step(pc-pr)*pow(ro,pc-pr)
    }
  }
  ro = dunif(-1,1)
  
  #Structure matrix for the third dimension (disease): 
  for (i in 1:(nGroups[3])){
    for (j in 1:(nGroups[3])){
      structure3[i,j] ~ dnorm(0,prec) 
    }
  }
  
  #Structure matrix for the fourth dimension (sex):
  for (i in 1:(nGroups[4])){
    for (j in 1:(nGroups[4])){
      structure4[i,j] ~ dnorm(0,prec)
    }
  }
  
  #Prior distribution for the random effects variability 
  prec <- pow(sdstruct,-2)
  sdstruct ~ dunif(0,100)
}

inits <- function() {
    list(mu = array(runif(Nper * Ndiseases * Nsex, -0.5, 0), dim = c(Nper, Ndiseases, Nsex)), ro = runif(1, 0, 1), structure3 = matrix(rnorm(Ndiseases * Ndiseases, 0, 1), nrow = Ndiseases), structure4 = matrix(rnorm(Nsex * 
        Nsex, 0, 1), nrow = Nsex), sdstruct = runif(1, 0, 10), gamma = runif(1, 0.6, 0.95), tS1 = array(rnorm(Nmuni * Nper * Ndiseases * Nsex, 0, 1), dim = c(Nsex, Ndiseases, Nper, Nmuni)))
}
param <- c("mu", "gamma", "ro", "structure3", "structure4", "sdstruct", "sSMR")

res1.2.3.4 <- pbugs(data = data, inits = inits, par = param, model = model1.2.3.4, n.iter = num.simu, n.burnin = num.burnin, bugs.seed = 1)
# Computing time
res1.2.3.4$exec_time
# Convergence assessment of identifiable parameters
aux <- c(grep("mu", dimnames(res1.2.3.4$summary)[[1]]), grep("gamma", dimnames(res1.2.3.4$summary)[[1]]), grep("ro", dimnames(res1.2.3.4$summary)[[1]]), grep("sdstruct", dimnames(res1.2.3.4$summary)[[1]]), 
    grep("sSMR", dimnames(res1.2.3.4$summary)[[1]]))
summary(res1.2.3.4$summary[aux, "Rhat"])
summary(res1.2.3.4$summary[aux, "n.eff"])
plot.deviance(res1.2.3.4)
```

##Spatial structure nested within period
```{r}
model12.2.3.4 = function(){
  for(k in 1:(nGroups[4])){#Sex
    for(j in 1:(nGroups[3])){#Disease
      for (p in 1:(nGroups[2])){#Period
        for(i in 1:(nGroups[1])){#Municipality
          Obs[i,p,j,k] ~ dpois(lambda[i,p,j,k])
          log(lambda[i,p,j,k]) <- log(Exp[i,p,j,k])+log.theta[i,p,j,k]
          log.theta[i,p,j,k] <- mu[p,j,k]+S1234[i,p,j,k]
          #Adds dependence between sexes
          S1234[i,p,j,k] <- inprod2(S123[i,p,j,],structure4[,k])
          #Adds dependence between diseases
          S123[i,p,j,k] <- inprod2(S12[i,p,,k],structure3[,j])
          #Adds dependence between periods
          S12[i,p,j,k] <- inprod2(tS1[k,j,,i],structure2[,p])

          sSMR[i,p,j,k] <- 100*exp(log.theta[i,p,j,k])
        }
        mu[p,j,k] ~ dflat()
        #Spatially dependent underlying random effects
        tS1[k,j,p,1:(nGroups[1])] ~ car.proper(ceros[],C[],adj[],num[],M[],1,gamma[p])
      }
    }
  }
  
  #Parameters for the PCAR distributions
  for(l in 1:(nGroups[1])){ceros[l] <- 0}
  gamma.inf <- min.bound(C[],adj[],num[],M[])
  gamma.sup <- max.bound(C[],adj[],num[],M[])
  for (p in 1:(nGroups[2])){
    gamma[p] ~ dunif(gamma.inf,gamma.sup)
  }
  
  #Structure matrix for the second dimension (period). Structure 2 is defined as the upper Cholesky matrix of the covariance matrix of an AR1 process.
  for (pc in 1:(nGroups[2])){
    #First Row
    structure2[1,pc] <- pow(ro,pc-1)
    #Rest of columns
    for (pr in 2:(nGroups[2])){
      structure2[pr,pc] <- step(pc-pr)*pow(ro,pc-pr)*pow((1-ro*ro),0.5)
    }
  }
  ro ~ dunif(-1,1)
  
  #Dimension 3 (Disease): 
  for (i in 1:(nGroups[3])){
    for (j in 1:(nGroups[3])){
      structure3[i,j] ~ dnorm(0,prec) 
    }
  }
  
  #Dimension 4 (Sex): 
  for (i in 1:(nGroups[4])){
    for (j in 1:(nGroups[4])){
      structure4[i,j] ~ dnorm(0,prec)
    }
  }
  
  #Prior distribution for the random effects variability
  prec <- pow(sdstruct,-2)
  sdstruct ~ dunif(0,100)
}  

inits <- function() {
    list(mu = array(runif(Nper * Ndiseases * Nsex, -0.5, 0), dim = c(Nper, Ndiseases, Nsex)), ro = runif(1, 0, 1), structure3 = matrix(rnorm(Ndiseases * Ndiseases, 0, 1), nrow = Ndiseases), structure4 = matrix(rnorm(Nsex * 
        Nsex, 0, 1), nrow = Nsex), sdstruct = runif(1, 0, 10), gamma = runif(Nper, 0.6, 0.95), tS1 = array(rnorm(Nmuni * Nper * Ndiseases * Nsex, 0, 1), dim = c(Nsex, Ndiseases, Nper, Nmuni)))
}
param <- c("mu", "gamma", "ro", "structure3", "structure4", "sdstruct", "sSMR")

res12.2.3.4 <- pbugs(data = data, inits = inits, par = param, model = model12.2.3.4, n.iter = num.simu, n.burnin = num.burnin, bugs.seed = 1)
# Computing time
res12.2.3.4$exec_time
# Convergence assessment of identifiable parameters
aux <- c(grep("mu", dimnames(res12.2.3.4$summary)[[1]]), grep("ro", dimnames(res12.2.3.4$summary)[[1]]), grep("sdstruct", dimnames(res12.2.3.4$summary)[[1]]), grep("sSMR", dimnames(res12.2.3.4$summary)[[1]]))
summary(res12.2.3.4$summary[aux, "Rhat"])
summary(res12.2.3.4$summary[aux, "n.eff"])
plot.deviance(res12.2.3.4)
```

```{r}
par(mfrow = c(2, 3))
plot(res12.2.3.4$sims.array[, 1, "gamma[1]"], type = "l")
lines(res12.2.3.4$sims.array[, 2, "gamma[1]"], col = 2)
lines(res12.2.3.4$sims.array[, 3, "gamma[1]"], col = 3)
plot(res12.2.3.4$sims.array[, 1, "gamma[2]"], type = "l")
lines(res12.2.3.4$sims.array[, 2, "gamma[2]"], col = 2)
lines(res12.2.3.4$sims.array[, 3, "gamma[2]"], col = 3)
plot(res12.2.3.4$sims.array[, 1, "gamma[3]"], type = "l")
lines(res12.2.3.4$sims.array[, 2, "gamma[3]"], col = 2)
lines(res12.2.3.4$sims.array[, 3, "gamma[3]"], col = 3)
plot(res12.2.3.4$sims.array[, 1, "gamma[4]"], type = "l")
lines(res12.2.3.4$sims.array[, 2, "gamma[4]"], col = 2)
lines(res12.2.3.4$sims.array[, 3, "gamma[4]"], col = 3)
plot(res12.2.3.4$sims.array[, 1, "gamma[5]"], type = "l")
lines(res12.2.3.4$sims.array[, 2, "gamma[5]"], col = 2)
lines(res12.2.3.4$sims.array[, 3, "gamma[5]"], col = 3)
plot(res12.2.3.4$sims.array[, 1, "ro"], type = "l", ylim = c(-1, 1))
lines(res12.2.3.4$sims.array[, 2, "ro"], col = 2)
lines(res12.2.3.4$sims.array[, 3, "ro"], col = 3)


```

##Spatial structure nested within disease
```{r, eval=FALSE}
model13.2.3.4 = function(){
  for(k in 1:(nGroups[4])){#Sex
    for(j in 1:(nGroups[3])){#Disease
      for (p in 1:(nGroups[2])){#Period
        for(i in 1:(nGroups[1])){#Municipality
          Obs[i,p,j,k] ~ dpois(lambda[i,p,j,k])
          log(lambda[i,p,j,k]) <- log(Exp[i,p,j,k])+log.theta[i,p,j,k]
          log.theta[i,p,j,k] <- mu[p,j,k]+S1234[i,p,j,k]
          #Adds dependence between sexes
          S1234[i,p,j,k] <- inprod2(S123[i,p,j,],structure4[,k])          
          #Adds dependence between diseases
          S123[i,p,j,k] <- inprod2(S12[i,p,,k],structure3[,j])
          #Adds dependence between periods
          S12[i,p,j,k] <- inprod2(tS1[k,j,,i],structure2[,p])
          
          sSMR[i,p,j,k] <- 100*exp(log.theta[i,p,j,k])

        }
        mu[p,j,k] ~ dflat()
        #Spatially dependent underlying random effects
        tS1[k,j,p,1:(nGroups[1])] ~ car.proper(ceros[],C[],adj[],num[],M[],1,gamma[j])
      }
    }
  }
  
  #Parameters for the PCAR distributions
  for(l in 1:(nGroups[1])){ceros[l] <- 0}
  gamma.inf <- min.bound(C[],adj[],num[],M[])
  gamma.sup <- max.bound(C[],adj[],num[],M[])
  for (j in 1:(nGroups[3])){
    gamma[j] ~ dunif(gamma.inf,gamma.sup)
  }
  
  #Structure matrix for the second dimension (period). Structure 2 is defined as the upper Cholesky matrix of the covariance matrix of an AR1 process.
  for (pc in 1:(nGroups[2])){
    #First Row
    structure2[1,pc] <- pow(ro,pc-1)*pow((1-ro*ro),-0.5)
    #Rest of columns
    for (pr in 2:(nGroups[2])){
      structure2[pr,pc] <- step(pc-pr)*pow(ro,pc-pr)
    }
  }
  ro ~ dunif(-1,1)
  
  #Dimension 3 (Disease): 
  for (i in 1:(nGroups[3])){
    for (j in 1:(nGroups[3])){
      structure3[i,j] ~ dnorm(0,prec) 
    }
  }
  
  #Dimension 4 (Sex): 
  for (i in 1:(nGroups[4])){
    for (j in 1:(nGroups[4])){
      structure4[i,j] ~ dnorm(0,prec)
    }
  }

  #Prior distribution for the random effects variability 
  prec <- pow(sdstruct,-2)
  sdstruct ~ dunif(0,100)
}  

inits <- function() {
    list(mu = array(runif(Nper * Ndiseases * Nsex, -0.5, 0), dim = c(Nper, Ndiseases, Nsex)), ro = runif(1, 0, 1), structure3 = matrix(rnorm(Ndiseases * Ndiseases, 0, 1), nrow = Ndiseases), structure4 = matrix(rnorm(Nsex * 
        Nsex, 0, 1), nrow = Nsex), sdstruct = runif(1, 0, 10), gamma = runif(Ndiseases, 0.6, 0.95), tS1 = array(rnorm(Nmuni * Nper * Ndiseases * Nsex, 0, 1), dim = c(Nsex, Ndiseases, Nper, Nmuni)))
}
param <- c("mu", "gamma", "ro", "structure3", "structure4", "sdstruct", "sSMR")

res13.2.3.4 <- pbugs(data = data, inits = inits, par = param, model = model13.2.3.4, n.iter = num.simu, n.burnin = num.burnin, bugs.seed = 1)
# Computing time
res13.2.3.4$exec_time
# Convergence assessment of identifiable parameters
aux <- c(grep("mu", dimnames(res13.2.3.4$summary)[[1]]), grep("ro", dimnames(res13.2.3.4$summary)[[1]]), grep("sdstruct", dimnames(res13.2.3.4$summary)[[1]]), grep("sSMR", dimnames(res13.2.3.4$summary)[[1]]))
summary(res13.2.3.4$summary[aux, "Rhat"])
summary(res13.2.3.4$summary[aux, "n.eff"])
plot.deviance(res13.2.3.4)
```

##Spatial structure nested within sex
```{r, eval=FALSE}
model14.2.3.4 = function(){
  for(k in 1:(nGroups[4])){#Sex
    for(j in 1:(nGroups[3])){#Disease
      for (p in 1:(nGroups[2])){#Period
        for(i in 1:(nGroups[1])){#Municipality
          Obs[i,p,j,k] ~ dpois(lambda[i,p,j,k])
          log(lambda[i,p,j,k]) <- log(Exp[i,p,j,k])+log.theta[i,p,j,k]
          log.theta[i,p,j,k] <- mu[p,j,k]+S1234[i,p,j,k]
          #Adds dependence between sexes
          S1234[i,p,j,k] <- inprod2(S123[i,p,j,],structure4[,k])
          #Adds dependence between diseases
          S123[i,p,j,k] <- inprod2(S12[i,p,,k],structure3[,j])
          #Adds dependence between periods
          S12[i,p,j,k] <- inprod2(tS1[k,j,,i],structure2[,p])

          sSMR[i,p,j,k] <- 100*exp(log.theta[i,p,j,k])
        }
        mu[p,j,k] ~ dflat()
        #Spatially dependent underlying random effects
        tS1[k,j,p,1:(nGroups[1])] ~ car.proper(ceros[],C[],adj[],num[],M[],1,gamma[k])
      }
    }
  }
  
  #Parameters for the PCAR distributions
  for(l in 1:(nGroups[1])){ceros[l] <- 0}
  gamma.inf <- min.bound(C[],adj[],num[],M[])
  gamma.sup <- max.bound(C[],adj[],num[],M[])
  for (k in 1:(nGroups[4])){
    gamma[k] ~ dunif(gamma.inf,gamma.sup)
  }
  
  #Structure matrix for the second dimension (period). Structure 2 is defined as the upper Cholesky matrix of the covariance matrix of an AR1 process.
  for (pc in 1:(nGroups[2])){
    #First Row
    structure2[1,pc] <- pow(ro,pc-1)*pow((1-ro*ro),-0.5)
    #Rest of columns
    for (pr in 2:(nGroups[2])){
      structure2[pr,pc] <- step(pc-pr)*pow(ro,pc-pr)
    }
  }
  ro ~ dunif(-1,1)
  
  #Dimension 3 (Disease): 
  for (i in 1:(nGroups[3])){
    for (j in 1:(nGroups[3])){
      structure3[i,j] ~ dnorm(0,prec) 
    }
  }
  
  #Dimension 4 (Sex): 
  for (i in 1:(nGroups[4])){
    for (j in 1:(nGroups[4])){
      structure4[i,j] ~ dnorm(0,prec)
    }
  }
  
  #Prior distribution for the random effects variability 
  prec <- pow(sdstruct,-2)
  sdstruct ~ dunif(0,100)
}  

inits <- function() {
    list(mu = array(runif(Nper * Ndiseases * Nsex, -0.5, 0), dim = c(Nper, Ndiseases, Nsex)), ro = runif(1, 0, 1), structure3 = matrix(rnorm(Ndiseases * Ndiseases, 0, 1), nrow = Ndiseases), structure4 = matrix(rnorm(Nsex * 
        Nsex, 0, 1), nrow = Nsex), sdstruct = runif(1, 0, 10), gamma = runif(Nsex, 0.6, 0.95), tS1 = array(rnorm(Nmuni * Nper * Ndiseases * Nsex, 0, 1), dim = c(Nsex, Ndiseases, Nper, Nmuni)))
}
param <- c("mu", "gamma", "ro", "structure3", "structure4", "sdstruct", "tS1", "sSMR")

res14.2.3.4 <- pbugs(data = data, inits = inits, par = param, model = model14.2.3.4, n.iter = num.simu, n.burnin = num.burnin, bugs.seed = 1)
# Computing time
res14.2.3.4$exec_time
# Convergence assessment of identifiable parameters
aux <- c(grep("mu", dimnames(res14.2.3.4$summary)[[1]]), grep("ro", dimnames(res14.2.3.4$summary)[[1]]), grep("sdstruct", dimnames(res14.2.3.4$summary)[[1]]), grep("sSMR", dimnames(res14.2.3.4$summary)[[1]]))
summary(res14.2.3.4$summary[aux, "Rhat"])
summary(res14.2.3.4$summary[aux, "n.eff"])
plot.deviance(res14.2.3.4)
```

##Temporal structure nested within disease
```{r, eval=FALSE}
model1.23.3.4 = function(){
  for(k in 1:(nGroups[4])){#Sex
    for(j in 1:(nGroups[3])){#Disease
      for (p in 1:(nGroups[2])){#Period
        for(i in 1:(nGroups[1])){#Municipality
          Obs[i,p,j,k] ~ dpois(lambda[i,p,j,k])
          log(lambda[i,p,j,k]) <- log(Exp[i,p,j,k])+log.theta[i,p,j,k]
          log.theta[i,p,j,k] <- mu[p,j,k]+S1234[i,p,j,k]
          #Adds dependence between sexes
          S1234[i,p,j,k] <- inprod2(S123[i,p,j,],structure4[,k])
          #Adds dependence between diseases
          S123[i,p,j,k] <- inprod2(S12[i,p,,k],structure3[,j])
          #Adds dependence between periods
          S12[i,p,j,k] <- inprod2(tS1[k,j,,i],structure2[,p,j])
          
          sSMR[i,p,j,k] <- 100*exp(log.theta[i,p,j,k])
        }
        mu[p,j,k] ~ dflat()
        #Spatially dependent underlying random effects
        tS1[k,j,p,1:(nGroups[1])] ~ car.proper(ceros[],C[],adj[],num[],M[],1,gamma)
      }
    }
  }
  
  #Parameters for the PCAR distributions
  for(l in 1:(nGroups[1])){ceros[l] <- 0}
  gamma.inf <- min.bound(C[],adj[],num[],M[])
  gamma.sup <- max.bound(C[],adj[],num[],M[])
  gamma ~ dunif(gamma.inf,gamma.sup)
  
  #Structure matrix for the second dimension (period). Structure 2 is defined as the upper Cholesky matrix of the covariance matrix of an AR1 process.
  for (j in 1:(nGroups[3])){
    for (pc in 1:(nGroups[2])){
      #First Row
      structure2[1,pc,j] <- pow(ro[j],pc-1)*pow((1-ro[j]*ro[j]),-0.5)
      #Rest of columns
      for (pr in 2:(nGroups[2])){
        structure2[pr,pc,j] <- step(pc-pr)*pow(ro[j],pc-pr)
      }
    }
    ro[j] ~ dunif(-1,1)
  }

  #Structure matrix for the third dimension (disease): 
  for (i in 1:(nGroups[3])){
    for (j in 1:(nGroups[3])){
      structure3[i,j] ~ dnorm(0,prec) 
    }
  }
  
  #Structure matrix for the third dimension (sex): 
  for (i in 1:(nGroups[4])){
    for (j in 1:(nGroups[4])){
      structure4[i,j] ~ dnorm(0,prec)
    }
  }
  
  #Prior distribution for the random effects variability 
  prec <- pow(sdstruct,-2)
  sdstruct ~ dunif(0,100)
}  

inits <- function() {
    list(mu = array(runif(Nper * Ndiseases * Nsex, -0.5, 0), dim = c(Nper, Ndiseases, Nsex)), ro = runif(Ndiseases, 0, 1), structure3 = matrix(rnorm(Ndiseases * Ndiseases, 0, 1), nrow = Ndiseases), structure4 = matrix(rnorm(Nsex * 
        Nsex, 0, 1), nrow = Nsex), sdstruct = runif(1, 0, 10), gamma = runif(1, 0.6, 0.95), tS1 = array(rnorm(Nmuni * Nper * Ndiseases * Nsex, 0, 1), dim = c(Nsex, Ndiseases, Nper, Nmuni)))
}
param <- c("mu", "gamma", "ro", "structure3", "structure4", "sdstruct", "tS1", "sSMR")
res1.23.3.4 <- pbugs(data = data, inits = inits, par = param, model = model1.23.3.4, n.iter = num.simu, n.burnin = num.burnin, bugs.seed = 1)
# Computing time
res1.23.3.4$exec_time
# Convergence assessment of identifiable parameters
aux <- c(grep("mu", dimnames(res1.23.3.4$summary)[[1]]), grep("gamma", dimnames(res1.23.3.4$summary)[[1]]), grep("sdstruct", dimnames(res1.23.3.4$summary)[[1]]), grep("sSMR", dimnames(res1.23.3.4$summary)[[1]]))
summary(res1.23.3.4$summary[aux, "Rhat"])
summary(res1.23.3.4$summary[aux, "n.eff"])
plot.deviance(res1.23.3.4)
```

##Temporal structure nested within sex
```{r, eval=FALSE}
model1.24.3.4 = function(){
  for(k in 1:(nGroups[4])){#Sex
    for(j in 1:(nGroups[3])){#Disease
      for (p in 1:(nGroups[2])){#Period
        for(i in 1:(nGroups[1])){#Municipality
          Obs[i,p,j,k] ~ dpois(lambda[i,p,j,k])
          log(lambda[i,p,j,k]) <- log(Exp[i,p,j,k])+log.theta[i,p,j,k]
          log.theta[i,p,j,k] <- mu[p,j,k]+S1234[i,p,j,k]
          #Adds dependence between sexes
          S1234[i,p,j,k] <- inprod2(S123[i,p,j,],structure4[,k])
          #Adds dependence between diseases
          S123[i,p,j,k] <- inprod2(S12[i,p,,k],structure3[,j])
          #Adds dependence between periods
          S12[i,p,j,k] <- inprod2(tS1[k,j,,i],structure2[,p,k])
          
          sSMR[i,p,j,k] <- 100*exp(log.theta[i,p,j,k])
        }
        mu[p,j,k] ~ dflat()
        #Spatially dependent underlying random effects
        tS1[k,j,p,1:(nGroups[1])] ~ car.proper(ceros[],C[],adj[],num[],M[],1,gamma)
      }
    }
  }
  
  #Parameters for the PCAR distributions
  for(l in 1:(nGroups[1])){ceros[l] <- 0}
  gamma.inf <- min.bound(C[],adj[],num[],M[])
  gamma.sup <- max.bound(C[],adj[],num[],M[])
  gamma ~ dunif(gamma.inf,gamma.sup)
  
  #Structure matrix for the second dimension (period). Structure 2 is defined as the upper Cholesky matrix of the covariance matrix of an AR1 process.
  for (k in 1:(nGroups[4])){
    for (pc in 1:(nGroups[2])){
      #First Row
      structure2[1,pc,k] <- pow(ro[k],pc-1)*pow((1-ro[k]*ro[k]),-0.5)
      #Rest of columns
      for (pr in 2:(nGroups[2])){
        structure2[pr,pc,k] <- step(pc-pr)*pow(ro[k],pc-pr)
      }
    }
    ro[k] ~ dunif(-1,1)
  }

  #Structure matrix for the third dimension (disease): 
  for (i in 1:(nGroups[3])){
    for (j in 1:(nGroups[3])){
      structure3[i,j] ~ dnorm(0,prec) 
    }
  }
  
  #Structure matrix for the third dimension (sex): 
  for (i in 1:(nGroups[4])){
    for (j in 1:(nGroups[4])){
      structure4[i,j] ~ dnorm(0,prec)
    }
  }
  
  #Prior distribution for the random effects variability 
  prec <- pow(sdstruct,-2)
  sdstruct ~ dunif(0,100)
}  

inits <- function() {
    list(mu = array(runif(Nper * Ndiseases * Nsex, -0.5, 0), dim = c(Nper, Ndiseases, Nsex)), ro = runif(Nsex, 0, 1), structure3 = matrix(rnorm(Ndiseases * Ndiseases, 0, 1), nrow = Ndiseases), structure4 = matrix(rnorm(Nsex * 
        Nsex, 0, 1), nrow = Nsex), sdstruct = runif(1, 0, 10), gamma = runif(1, 0.6, 0.95), tS1 = array(rnorm(Nmuni * Nper * Ndiseases * Nsex, 0, 1), dim = c(Nsex, Ndiseases, Nper, Nmuni)))
}
param <- c("mu", "gamma", "ro", "structure3", "structure4", "sdstruct", "sSMR")

res1.24.3.4 <- pbugs(data = data, inits = inits, par = param, model = model1.24.3.4, n.iter = num.simu, n.burnin = num.burnin, bugs.seed = 1)
# Computing time
res1.24.3.4$exec_time
# Convergence assessment of identifiable parameters
aux <- c(grep("mu", dimnames(res1.24.3.4$summary)[[1]]), grep("gamma", dimnames(res1.24.3.4$summary)[[1]]), grep("sdstruct", dimnames(res1.24.3.4$summary)[[1]]), grep("sSMR", dimnames(res1.24.3.4$summary)[[1]]))
summary(res1.24.3.4$summary[aux, "Rhat"])
summary(res1.24.3.4$summary[aux, "n.eff"])
plot.deviance(res1.24.3.4)
```

##Disease structure nested within Period
```{r,eval=FALSE}
model1.2.32.4 = function(){
  for(k in 1:(nGroups[4])){#Sex
    for(j in 1:(nGroups[3])){#Disease
      for (p in 1:(nGroups[2])){#Period
        for(i in 1:(nGroups[1])){#Municipality
          Obs[i,p,j,k] ~ dpois(lambda[i,p,j,k])
          log(lambda[i,p,j,k]) <- log(Exp[i,p,j,k])+log.theta[i,p,j,k]
          log.theta[i,p,j,k] <- mu[p,j,k]+S1234[i,p,j,k]
          #Adds dependence between sexes
          S1234[i,p,j,k] <- inprod2(S123[i,p,j,],structure4[,k])
          #Adds dependence between diseases
          S123[i,p,j,k] <- inprod2(S12[i,p,,k],structure3[,j,p])
          #Adds dependence between periods
          S12[i,p,j,k] <- inprod2(tS1[k,j,,i],structure2[,p])
          
          sSMR[i,p,j,k] <- 100*exp(log.theta[i,p,j,k])
        }
        mu[p,j,k] ~ dflat()
        #Spatially dependent underlying random effects
        tS1[k,j,p,1:(nGroups[1])] ~ car.proper(ceros[],C[],adj[],num[],M[],1,gamma)
      }
    }
  }

  #Parameters for the PCAR distributions
  for(l in 1:(nGroups[1])){ceros[l] <- 0}
  gamma.inf <- min.bound(C[],adj[],num[],M[])
  gamma.sup <- max.bound(C[],adj[],num[],M[])
  gamma ~ dunif(gamma.inf,gamma.sup)
  
  #Structure matrix for the second dimension (period). Structure 2 is defined as the upper Cholesky matrix of the covariance matrix of an AR1 process.
  for (pc in 1:(nGroups[2])){
    #First Row
    structure2[1,pc] <- pow(ro,pc-1)*pow((1-ro*ro),-0.5)
    #Rest of columns
    for (pr in 2:(nGroups[2])){
      structure2[pr,pc] <- step(pc-pr)*pow(ro,pc-pr)
    }
  }
  ro ~ dunif(-1,1)
  
  #Structure matrix for the third dimension (disease): 
  for (p in 1:(nGroups[2])){
    for (i in 1:(nGroups[3])){
      for (j in 1:(nGroups[3])){
        structure3[i,j,p] ~ dnorm(0,prec) 
      }
    }
  }

  #Structure matrix for the third dimension (sex): 
  for (i in 1:(nGroups[4])){
    for (j in 1:(nGroups[4])){
      structure4[i,j] ~ dnorm(0,prec)
    }
  }
  
  #Prior distribution for the random effects variability 
  prec <- pow(sdstruct,-2)
  sdstruct ~ dunif(0,100)
}  

inits <- function() {
    list(mu = array(runif(Nper * Ndiseases * Nsex, -0.5, 0), dim = c(Nper, Ndiseases, Nsex)), ro = runif(1, 0, 1), structure3 = array(rnorm(Ndiseases * Ndiseases * Nper, 0, 1), dim = c(Ndiseases, Ndiseases, 
        Nper)), structure4 = matrix(rnorm(Nsex * Nsex, 0, 1), nrow = Nsex), sdstruct = runif(1, 0, 10), gamma = runif(1, 0.6, 0.95), tS1 = array(rnorm(Nmuni * Nper * Ndiseases * Nsex, 0, 1), dim = c(Nsex, 
        Ndiseases, Nper, Nmuni)))
}
param <- c("mu", "gamma", "ro", "structure3", "structure4", "sdstruct", "sSMR")

res1.2.32.4 <- pbugs(data = data, inits = inits, par = param, model = model1.2.32.4, n.iter = num.simu, n.burnin = num.burnin, bugs.seed = 1)
# Computing time
res1.2.32.4$exec_time
# Convergence assessment of identifiable parameters
aux <- c(grep("mu", dimnames(res1.2.32.4$summary)[[1]]), grep("gamma", dimnames(res1.2.32.4$summary)[[1]]), grep("ro", dimnames(res1.2.32.4$summary)[[1]]), grep("sdstruct", dimnames(res1.2.32.4$summary)[[1]]), 
    grep("sSMR", dimnames(res1.2.32.4$summary)[[1]]))
summary(res1.2.32.4$summary[aux, "Rhat"])
summary(res1.2.32.4$summary[aux, "n.eff"])
plot.deviance(res1.2.32.4)
```

##Disease structure nested within sex
```{r,eval=FALSE}
model1.2.34.4 = function(){
  for(k in 1:(nGroups[4])){#Sex
    for(j in 1:(nGroups[3])){#Disease
      for (p in 1:(nGroups[2])){#Period
        for(i in 1:(nGroups[1])){#Municipality
          Obs[i,p,j,k] ~ dpois(lambda[i,p,j,k])
          log(lambda[i,p,j,k]) <- log(Exp[i,p,j,k])+log.theta[i,p,j,k]
          log.theta[i,p,j,k] <- mu[p,j,k]+S1234[i,p,j,k]
          #Adds dependence between sexes
          S1234[i,p,j,k] <- inprod2(S123[i,p,j,],structure4[,k])
          #Adds dependence between diseases
          S123[i,p,j,k] <- inprod2(S12[i,p,,k],structure3[,j,k])
          #Adds dependence between periods
          S12[i,p,j,k] <- inprod2(tS1[k,j,,i],structure2[,p])
          
          sSMR[i,p,j,k] <- 100*exp(log.theta[i,p,j,k])
        }
        mu[p,j,k] ~ dflat()

        #Spatially dependent underlying random effects
        tS1[k,j,p,1:(nGroups[1])] ~ car.proper(ceros[],C[],adj[],num[],M[],1,gamma)
      }
    }
  }
  
  #Parameters for the PCAR distributions
  for(l in 1:(nGroups[1])){ceros[l] <- 0}
  gamma.inf <- min.bound(C[],adj[],num[],M[])
  gamma.sup <- max.bound(C[],adj[],num[],M[])
  gamma ~ dunif(gamma.inf,gamma.sup)
  
  #Structure matrix for the second dimension (period). Structure 2 is defined as the upper Cholesky matrix of the covariance matrix of an AR1 process.
  for (pc in 1:(nGroups[2])){
    #First Row
    structure2[1,pc] <- pow(ro,pc-1)*pow((1-ro*ro),-0.5)
    #Rest of columns
    for (pr in 2:(nGroups[2])){
      structure2[pr,pc] <- step(pc-pr)*pow(ro,pc-pr)
    }
  }
  ro~dunif(-1,1)
  
  #Structure matrix for the fourth dimension (disease):
  for (k in 1:(nGroups[4])){
    for (i in 1:(nGroups[3])){
      for (j in 1:(nGroups[3])){
        structure3[i,j,k] ~ dnorm(0,prec) 
      }
    }
  }
  
  #Structure matrix for the fourth dimension (sex):
  for (i in 1:(nGroups[4])){
    for (j in 1:(nGroups[4])){
      structure4[i,j] ~ dnorm(0,prec)
    }
  }

  #Prior distribution for the random effects variability 
  prec <- pow(sdstruct,-2)
  sdstruct ~ dunif(0,100)
}  

inits <- function() {
    list(mu = array(runif(Nper * Ndiseases * Nsex, -0.5, 0), dim = c(Nper, Ndiseases, Nsex)), ro = runif(1, 0, 1), structure3 = array(rnorm(Ndiseases * Ndiseases * Nsex, 0, 1), dim = c(Ndiseases, Ndiseases, 
        Nsex)), structure4 = matrix(rnorm(Nsex * Nsex, 0, 1), nrow = Nsex), sdstruct = runif(1, 0, 10), gamma = runif(1, 0.6, 0.95), tS1 = array(rnorm(Nmuni * Nper * Ndiseases * Nsex, 0, 1), dim = c(Nsex, 
        Ndiseases, Nper, Nmuni)))
}
param <- c("mu", "gamma", "ro", "structure3", "structure4", "sdstruct", "sSMR")

res1.2.34.4 <- pbugs(data = data, inits = inits, par = param, model = model1.2.34.4, n.iter = num.simu, n.burnin = num.burnin, bugs.seed = 1)
# Computing time
res1.2.34.4$exec_time
# Convergence assessment of identifiable parameters
aux <- c(grep("mu", dimnames(res1.2.34.4$summary)[[1]]), grep("gamma", dimnames(res1.2.34.4$summary)[[1]]), grep("ro", dimnames(res1.2.34.4$summary)[[1]]), grep("sdstruct", dimnames(res1.2.34.4$summary)[[1]]), 
    grep("sSMR", dimnames(res1.2.34.4$summary)[[1]]))
summary(res1.2.34.4$summary[aux, "Rhat"])
summary(res1.2.34.4$summary[aux, "n.eff"])
plot.deviance(res1.2.34.4)
```

##Sex structure nested within period
```{r, eval=FALSE}
model1.2.3.42 = function(){
  for(k in 1:(nGroups[4])){#Sex
    for(j in 1:(nGroups[3])){#Disease
      for (p in 1:(nGroups[2])){#Period
        for(i in 1:(nGroups[1])){#Municipality
          Obs[i,p,j,k] ~ dpois(lambda[i,p,j,k])
          log(lambda[i,p,j,k]) <- log(Exp[i,p,j,k])+log.theta[i,p,j,k]
          log.theta[i,p,j,k] <- mu[p,j,k]+S1234[i,p,j,k]
          #Adds dependence between sexes
          S1234[i,p,j,k] <- inprod2(S123[i,p,j,],structure4[,k,p])
          #Adds dependence between disease
          S123[i,p,j,k] <- inprod2(S12[i,p,,k],structure3[,j])
          #Adds dependence between period
          S12[i,p,j,k] <- inprod2(tS1[k,j,,i],structure2[,p])
          
          sSMR[i,p,j,k] <- 100*exp(log.theta[i,p,j,k])
        }
        mu[p,j,k] ~ dflat()
        #Spatially dependent underlying random effects
        tS1[k,j,p,1:(nGroups[1])] ~ car.proper(ceros[],C[],adj[],num[],M[],1,gamma)
      }
    }
  }
  
  #Parameters for the PCAR distributions
  for(l in 1:(nGroups[1])){ceros[l] <- 0}
  gamma.inf <- min.bound(C[],adj[],num[],M[])
  gamma.sup <- max.bound(C[],adj[],num[],M[])
  gamma ~ dunif(gamma.inf,gamma.sup)
  
  #Structure matrix for the second dimension (period). Structure 2 is defined as the upper Cholesky matrix of the covariance matrix of an AR1 process.
  for (pc in 1:(nGroups[2])){
    #First Row
    structure2[1,pc] <- pow(ro,pc-1)*pow((1-ro*ro),-0.5)
    #Rest of columns
    for (pr in 2:(nGroups[2])){
      structure2[pr,pc] <- step(pc-pr)*pow(ro,pc-pr)
    }
  }
  ro~dunif(-1,1)
  
  #Structure matrix for the third dimension (disease): 
  for (i in 1:(nGroups[3])){
    for (j in 1:(nGroups[3])){
      structure3[i,j] ~ dnorm(0,prec) 
    }
  }
  
  #Structure matrix for the third dimension (sex): 
  for (k in 1:(nGroups[2])){ 
    for (i in 1:(nGroups[4])){
      for (j in 1:(nGroups[4])){
        structure4[i,j,k] ~ dnorm(0,prec)
      }
    }
  }

  #Prior distribution for the random effects variability 
  prec <- pow(sdstruct,-2)
  sdstruct ~ dunif(0,100)
}  

inits <- function() {
    list(mu = array(runif(Nper * Ndiseases * Nsex, -0.5, 0), dim = c(Nper, Ndiseases, Nsex)), ro = runif(1, 0, 1), structure3 = matrix(rnorm(Ndiseases * Ndiseases, 0, 1), nrow = Ndiseases), structure4 = array(rnorm(Nsex * 
        Nsex * Nper, 0, 1), dim = c(Nsex, Nsex, Nper)), sdstruct = runif(1, 0, 10), tS1 = array(rnorm(Nmuni * Nper * Ndiseases * Nsex, 0, 1), dim = c(Nsex, Ndiseases, Nper, Nmuni)))
}
param <- c("mu", "gamma", "ro", "structure3", "structure4", "sdstruct", "sSMR")

res1.2.3.42 <- pbugs(data = data, inits = inits, par = param, model = model1.2.3.42, n.iter = num.simu, n.burnin = num.burnin, n.chains = 3)
# Computing time
res1.2.3.42$exec_time
# Convergence assessment of identifiable parameters
aux <- c(grep("mu", dimnames(res1.2.3.42$summary)[[1]]), grep("gamma", dimnames(res1.2.3.42$summary)[[1]]), grep("ro", dimnames(res1.2.3.42$summary)[[1]]), grep("sdstruct", dimnames(res1.2.3.42$summary)[[1]]), 
    grep("sSMR", dimnames(res1.2.3.42$summary)[[1]]))
summary(res1.2.3.42$summary[aux, "Rhat"])
summary(res1.2.3.42$summary[aux, "n.eff"])
plot.deviance(res1.2.3.42)
```

##Sex structure nested within disease
```{r, eval=FALSE}
model1.2.3.43 = function(){
  for(k in 1:(nGroups[4])){#Sex
    for(j in 1:(nGroups[3])){#Disease
      for (p in 1:(nGroups[2])){#Period
        for(i in 1:(nGroups[1])){#Municipality
          Obs[i,p,j,k] ~ dpois(lambda[i,p,j,k])
          log(lambda[i,p,j,k]) <- log(Exp[i,p,j,k])+log.theta[i,p,j,k]
          log.theta[i,p,j,k] <- mu[p,j,k]+S1234[i,p,j,k]
          #Adds dependence between sexes
          S1234[i,p,j,k] <- inprod2(S123[i,p,j,],structure4[,k,j])
          #Adds dependence between diseases
          S123[i,p,j,k] <- inprod2(S12[i,p,,k],structure3[,j])
          #Adds dependence between periods
          S12[i,p,j,k] <- inprod2(tS1[k,j,,i],structure2[,p])
          
          sSMR[i,p,j,k] <- 100*exp(log.theta[i,p,j,k])          
        }
        mu[p,j,k] ~ dflat()
        #Spatially dependent underlying random effects
        tS1[k,j,p,1:(nGroups[1])] ~ car.proper(ceros[],C[],adj[],num[],M[],1,gamma)
      }
    }
  }

  #Parameters for the PCAR distributions
  for(l in 1:(nGroups[1])){ceros[l] <- 0}
  gamma.inf <- min.bound(C[],adj[],num[],M[])
  gamma.sup <- max.bound(C[],adj[],num[],M[])
  gamma ~ dunif(gamma.inf,gamma.sup)
  
  #Structure matrix for the second dimension (period). Structure 2 is defined as the upper Cholesky matrix of the covariance matrix of an AR1 process.
  for (pc in 1:(nGroups[2])){
    #First Row
    structure2[1,pc] <- pow(ro,pc-1)*pow((1-ro*ro),-0.5)
    #Rest of columns
    for (pr in 2:(nGroups[2])){
      structure2[pr,pc] <- step(pc-pr)*pow(ro,pc-pr)
    }
  }
  ro ~ dunif(-1,1)
  
  #Structure matrix for the third dimension (disease): 
  for (i in 1:(nGroups[3])){
    for (j in 1:(nGroups[3])){
      structure3[i,j] ~ dnorm(0,prec) 
    }
  }
  
  #Structure matrix for the third dimension (sex): 
  for (k in 1:(nGroups[3])){ 
    for (i in 1:(nGroups[4])){
      for (j in 1:(nGroups[4])){
        structure4[i,j,k] ~ dnorm(0,prec)
      }
    }
  }
  
  #Prior distribution for the random effects variability 
  prec <- pow(sdstruct,-2)
  sdstruct ~ dunif(0,100)
}  

inits <- function() {
    list(mu = array(runif(Nper * Ndiseases * Nsex, -0.5, 0), dim = c(Nper, Ndiseases, Nsex)), ro = runif(1, 0, 1), structure3 = matrix(rnorm(Ndiseases * Ndiseases, 0, 1), nrow = Ndiseases), structure4 = array(rnorm(Nsex * 
        Nsex * Ndiseases, 0, 1), dim = c(Nsex, Nsex, Ndiseases)), sdstruct = runif(1, 0, 10), tS1 = array(rnorm(Nmuni * Nper * Ndiseases * Nsex, 0, 1), dim = c(Nsex, Ndiseases, Nper, Nmuni)))
}
param <- c("mu", "gamma", "ro", "structure3", "structure4", "sdstruct", "sSMR")

res1.2.3.43 <- pbugs(data = data, inits = inits, par = param, model = model1.2.3.43, n.iter = num.simu, n.burnin = num.burnin, bugs.seed = 1)
# Computing time
res1.2.3.43$exec_time
# Convergence assessment of identifiable parameters
aux <- c(grep("mu", dimnames(res1.2.3.43$summary)[[1]]), grep("gamma", dimnames(res1.2.3.43$summary)[[1]]), grep("ro", dimnames(res1.2.3.43$summary)[[1]]), grep("sdstruct", dimnames(res1.2.3.43$summary)[[1]]), 
    grep("sSMR", dimnames(res1.2.3.43$summary)[[1]]))
summary(res1.2.3.43$summary[aux, "Rhat"])
summary(res1.2.3.43$summary[aux, "n.eff"])
plot.deviance(res1.2.3.43)
```

