---
title: "Example 4.5"
subtitle: "Disease mapping: from foundations to multidimensional modeling"
author: Martinez-Beneito M.A. and Botella-Rocamora P.
output: pdf_document
editor_options: 
  chunk_output_type: console
---
  
This document reproduces the analysis made at Example 4.5 of the book: "Disease mapping: from foundations to multidimensional modeling" by Martinez-Beneito M.A. and Botella-Rocamora P., published by CRC press in 2019. You can watch the analysis made with full detail at this pdf document, or even execute it if you want with the material available at <https://github.com/MigueBeneito/DMBook>. Anyway, this pdf file should be enough for following most of the details of the analysis made for this example.

The statistical analysis below has been run in `R`, by additionally using the library `Rmarkdown`, so be sure that you have this software installed if you want to reproduce by yourself the content of this document. In that case we advise you to download first the annex material at <https://github.com/MigueBeneito/DMBook>, open with `Rstudio` the corresponding `.Rproj` file that you will find at the folder corresponding to this example and compile the corresponding `.Rmd` document. This will allow you to reproduce the whole statistical analysis below.
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, width = 80, tidy = TRUE, tidy.opts = list(width.cutoff = 70))
set.seed(1)
```

This document has been executed with real data that are not provided in order to preserve their confidentiality. Slightly modified data are provided instead, as described in Chapter 1 of the book. Thus, when reproducing this document you will not obtain exactly the same results, although they should be very close to those shown here.

##Libraries and data loading
```{r libraries and data loading, warning=FALSE, message=FALSE}
#Libraries loading
#-----------------
if(!require(RColorBrewer)){install.packages("RColorBrewer");library(RColorBrewer)}
if(!require(rgdal)){install.packages("rgdal");library(rgdal)}
if(!require(INLA)){install.packages("INLA",repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE);library(INLA)}
if(!require(pbugs)){
    if(!require(devtools)){
        install.packages("devtools");devtools::install_github("fisabio/pbugs")
    }
    else{
        install_github("fisabio/pbugs")
    }
}

#Data loading
#------------
#For reproducing the document, the following line should be changed to load("../Data/ObsOral-mod.Rdata") since that file contains the modified data making it possible to reproduce this document.
load("../Data/ObsOral.Rdata")
#load("../Data/ObsOral-mod.Rdata")
load("../Data/ExpOral.Rdata")
load("../Data/VR.Rdata")
```
##`R` function for calculating the DIC criterion of the models fitted
The function below computes the DIC criterion for disease mapping models fitted with `WinBUGS`. It returns DIC values comparable to those reported by `INLA`, in contrast to `WinBUGS`. See annex material for Example 4.3.
```{r}
# Arguments: Simu.sSMRs: matrix of dimensions n.IterXn.Units where n.Iter are the number of MCMC iterations saved and n.Units the number of spatial units in the analysis. You will typically find this
# as a submatrix of the sims.matrix element of any bugs object. O: Vector of length n.Units with the observed deaths per spatial unit. E: Vector of length n.Units with the expected deaths per spatial
# unit.
DICPoisson = function(Simu.sSMRs, O, E) {
    mu = t(apply(Simu.sSMRs/100, 1, function(x) {
        x * E
    }))
    D = apply(mu, 1, function(x) {
        -2 * sum(O * log(x) - x - lfactorial(O))
    })
    Dmean = mean(D)
    mumean = apply(Simu.sSMRs/100, 2, mean) * E
    DinMean = -2 * sum(O * log(mumean) - mumean - lfactorial(O))
    # if(save==TRUE){return(c(Dmedia,Dmedia-DenMedia,2*Dmedia-DenMedia))}
    cat("D=", Dmean, "pD=", Dmean - DinMean, "DIC=", 2 * Dmean - DinMean, "\n")
}
```

##INLA implementation of the BYM model
```{r}
sdunif = "expression:
  logdens = -log_precision/2;
  return(logdens)"

data = data.frame(O = Obs.muni, E = Exp.muni, id.node = 1:540)
# See the annex material of Example 3.6 for details on how the VR.graph file is generated
form = O ~ f(id.node, model = "bym", hyper = list(prec.spatial = list(prior = sdunif), prec.unstruct = list(prior = sdunif)), graph = "../Data/VR.graph")
resul.BYM.inla = inla(form, family = "poisson", data = data, E = E, control.compute = list(dic = TRUE))
# Computing time
resul.BYM.inla$cpu.used
```

##INLA implementation of some alternative spatial models
```{r}
# PCAR distributed random effects
formProper = O ~ f(id.node, model = "besagproper", hyper = list(prec = list(prior = sdunif)), graph = "../Data/VR.graph")
resul.PCAR.inla <- inla(formProper, family = "poisson", data = data, E = E, control.compute = list(dic = TRUE))
# Computing time
resul.PCAR.inla$cpu.used

# HCAR distributed random effects Cmatrix exportation as required by INLA
write.table(cbind(rep(1:540, VR.wb$num), VR.wb$adj, rep(1, length(VR.wb$adj))), "../Data/Cmat.inla", row.names = FALSE, col.names = FALSE)
formGeneric = O ~ f(id.node, model = "generic1", hyper = list(prec = list(prior = sdunif)), Cmatrix = "../Data/Cmat.inla")
resul.HCAR.inla <- inla(formGeneric, family = "poisson", data = data, E = E, control.compute = list(dic = TRUE))
# Computing time
resul.HCAR.inla$cpu.used
```

##Proper CAR model fit with WinBUGS
```{r}
ModelProperCAR = function(){
 for(i in 1:n){
  O[i] ~ dpois(lambda[i])
  log(lambda[i]) <- log(E[i])+log.theta[i]
  log.theta[i] <- sp[i]
  sSMR[i] <- 100*exp(log.theta[i])
  P.sSMR[i] <- step(sSMR[i]-100)
 }
 
 #The proper CAR "piece"
 sp[1:n] ~ car.proper(mu.sp[], C[], adj[], num[], M[], tau, rho)
 for(i in 1:n){mu.sp[i] <- mu}
 
 #Prior distributions
 rho ~ dunif(rho.low,rho.up)
 rho.low <- min.bound(C[], adj[], num[], M[])
 rho.up <- max.bound(C[], adj[], num[], M[])
 tau <- pow(sd.sp,-2)
 sd.sp ~ dunif(0,10)
 mu ~ dflat()
}

data <- list(O = Obs.muni, E = Exp.muni, n = 540, num = VR.wb$num, adj = VR.wb$adj, M = 1/VR.wb$num, C = rep(1/VR.wb$num, VR.wb$num))
inits <- function() {
    list(sp = rnorm(540), sd.sp = runif(1), mu = rnorm(1))
}
param <- c("sSMR", "mu", "sd.sp", "rho")
# BEWARE: The following line could take long to run.
time.ProperCAR <- system.time(ResulProperCAR <- pbugs(data = data, inits = inits, param = param, model.file = ModelProperCAR, n.iter = 150000, n.burnin = 20000, bugs.seed = 1, DIC = FALSE))
# Computing time
ResulProperCAR$exec_time
# Result summaries
summary(ResulProperCAR$summary[, "Rhat"])
summary(ResulProperCAR$summary[, "n.eff"])
round(ResulProperCAR$summary[c("mu", "sd.sp", "rho"), ], 1)
# DIC
DICPoisson(ResulProperCAR$sims.matrix[, grep("sSMR", dimnames(ResulProperCAR$sims.matrix)[[2]])], Obs.muni, Exp.muni)
```

##BYM model fit with WinBUGS
```{r}
ModelBYM = function(){
 for(i in 1:n){
  O[i] ~ dpois(lambda[i])
  log(lambda[i]) <- log(E[i])+log.theta[i]
  log.theta[i] <- mu+sp[i]+het[i]
  het[i] ~ dnorm(0,tau.het)
  sSMR[i] <- 100*exp(log.theta[i])
  P.sSMR[i] <- step(sSMR[i]-100)
 }

 #The intrinsic CAR "piece"
 sp[1:n] ~ car.normal(adj[], weights[], num[], tau.sp)
 for(i in 1:nadj){weights[i] <- 1}

 #Prior distributions
 tau.het <- pow(sd.het,-2)
 sd.het ~ dunif(0,10)
 tau.sp <- pow(sd.sp,-2)
 sd.sp ~ dunif(0,10)
 mu ~ dflat()
}

data <- list(O = Obs.muni, E = Exp.muni, n = 540, num = VR.wb$num, adj = VR.wb$adj, nadj = length(VR.wb$adj))
inits <- function() {
    list(sp = rnorm(540), het = rnorm(540), sd.sp = runif(1), sd.het = runif(1), mu = rnorm(1))
}
param <- c("sSMR", "mu", "sd.sp", "sd.het")
# BEWARE: The following line could take long to run.
ResulBYM <- pbugs(data = data, inits = inits, param = param, model.file = ModelBYM, n.iter = 150000, n.burnin = 20000, bugs.seed = 1, DIC = FALSE)
# Computing time
ResulBYM$exec_time
# Result summaries
summary(ResulBYM$summary[, "Rhat"])
summary(ResulBYM$summary[, "n.eff"])
round(ResulBYM$summary[c("mu", "sd.sp", "sd.het"), ], 1)
# DIC
DICPoisson(ResulBYM$sims.matrix[, grep("sSMR", dimnames(ResulBYM$sims.matrix)[[2]])], Obs.muni, Exp.muni)
# Highest sSMR
max(ResulBYM$mean$sSMR)
```

##Choropleth maps and some summaries
```{r}
ColorSMR <- brewer.pal(7, "BrBG")[7:1]
ColorProb <- brewer.pal(7, "RdYlGn")[7:1]
par(mfrow = c(1, 2))
par(mar = c(1, 1, 2, 1) + 0.1)
plot(VR.cart, col = ColorSMR[as.numeric(cut(ResulBYM$mean$sSMR, 100 * c(-0.1, 1/1.5, 1/1.25, 1/1.1, 1.1, 1.25, 1.5, 100)))])
title("BYM model", cex = 0.75)
legend(x = "bottomright", fill = ColorSMR[7:1], legend = c(">150", "125-150", "110-125", "91-110", "80-91", "66-80", "<66"), cex = 0.65, inset = 0.03, title = "sSMR")
plot(VR.cart, col = ColorProb[as.numeric(cut(apply(ResulBYM$sims.list$sSMR, 2, function(x) {
    mean(x > 100)
}), c(-0.1, 0.05, 0.1, 0.2, 0.8, 0.9, 0.95, 1.1)))])
title("BYM model", cex = 0.75)
legend(x = "bottomright", fill = ColorProb[7:1], legend = c(">0.95", "0.9-0.95", "0.8-0.9", "0.2-0.8", "0.1-0.2", "0.05-0.1", "<0.05"), cex = 0.65, inset = 0.03, title = "Probability")

# Correlations between spatial patterns of the sSMRs
cor(cbind(ResulProperCAR$mean$sSMR, ResulBYM$mean$sSMR))
# standard deviations of the sSMRs for each spatial pattern
sd(ResulProperCAR$mean$sSMR)
sd(ResulBYM$mean$sSMR)
```

##Standard deviations and correlations between different sSMRs patterns
```{r}
#standard deviation for the spatial patterns fitted
100 * sd(resul.BYM.inla$summary.fitted.values[, 1])
100 * sd(resul.PCAR.inla$summary.fitted.values[, 1])
100 * sd(resul.HCAR.inla$summary.fitted.values[, 1])
sd(ResulBYM$mean$sSMR)
sd(ResulProperCAR$mean$sSMR)

#Correlations between patterns
round(cor(cbind(resul.BYM.inla$summary.fitted.values[, 1], resul.PCAR.inla$summary.fitted.values[, 1], resul.HCAR.inla$summary.fitted.values[, 1], ResulBYM$mean$sSMR, ResulProperCAR$mean$sSMR)), 4)
```

